---
title: "Deep dive: multi-species IPMs and the forestIPM roadmap"
---

This chapter extends the single-species quick start in @sec-quickstart to cover multi-species IPMs, time-varying climate, posterior uncertainty propagation, and plot-level random effects.
The package API is identical to what is described in the guide — only the argument values change.

# Multi-species IPMs {#sec-deepdive}

Running a multi-species IPM requires only one change to the quick-start workflow: include multiple `species_id` values in the `stand()` call.
Here we build a two-species stand combining *Abies balsamea* (ABIBAL) and *Acer rubrum* (ACERUB), a common mixedwood assemblage in eastern North America.

## Building a multi-species stand

```{r multisp-setup}
library(forestIPM)

# Two-species stand on a 400 m² plot
s_multi <- stand(
  data.frame(
    size_mm    = c(150, 200, 180, 350, 280),
    species_id = c("ABIBAL", "ABIBAL", "ACERUB", "ACERUB", "ABIBAL"),
    plot_size  = 400
  )
)
s_multi
```

`stand()` accepts any number of species — just mix them in the `species_id` column.
The column name is flexible: `species_id`, `species`, or `sp` are all accepted.
Use `summary(s_multi)` to confirm the per-species tree counts and size ranges.

## Handling species with missing parameters

When the stand includes a species that is not (yet) in `supported_species()`, the `on_missing` argument of `species_model()` controls what happens:

```{r on-missing, eval=FALSE}
# Default: error with a closest-match suggestion
mod <- species_model(s_multi, on_missing = "error")

# Silently drop the unknown species; model runs for remaining species
mod <- species_model(s_multi, on_missing = "drop")

# Keep the species as a static competitor (its Nvec never updates)
mod <- species_model(s_multi, on_missing = "static")
```

`"static"` is useful when you want to fix a competitor's size distribution throughout the simulation — for example, to isolate the effect of a single focal species while holding its competitors constant.

## Running the multi-species projection

`species_model()` detects both species automatically and loads their respective fitted models.
Interspecific competition between them is captured internally via the basal area competition terms.

```{r multisp-run, eval=FALSE}
mod_multi  <- species_model(s_multi)
pars_multi <- parameters(mod_multi, draw = "mean")
env_multi  <- env_condition(MAT = 8, MAP = 1200)
ctrl_multi <- control(years = 100, compute_lambda = TRUE, progress = FALSE)

proj_multi <- project(mod_multi, pars_multi, s_multi, env_multi, ctrl_multi)
```

```{r multisp-plot, eval=FALSE}
#| label: fig-multisp
#| fig-cap: "Lambda trajectories for ABIBAL (balsam fir) and ACERUB (red maple) in a mixed stand. Species with a competitive advantage under these conditions maintain $\\lambda > 1$ longer."
plot(proj_multi, type = "lambda")
```

The `$summary` tibble of the returned `ipm_projection` contains one row per species per stored timestep, so you can filter and plot directly:

```{r multisp-filter, eval=FALSE}
library(dplyr)
proj_multi$summary |>
  filter(species_id == "ABIBAL") |>
  select(timestep, lambda, n_trees)
```

## Exploring species availability

If you are unsure which species IDs are available, use `supported_species()`:

```{r supported}
supported_species()
```

The tibble also lists the fitted climate envelope (`MAT_min`, `MAT_max`, `MAP_min`, `MAP_max`) for each species — useful for choosing realistic `env_condition()` values and interpreting out-of-range warnings.

# Time-varying climate {#sec-time-varying}

A single numeric value in `env_condition()` represents a static climate.
To simulate climate change, pass a `function(t)` for `MAT` or `MAP`; the function receives the current simulation year `t` and returns the corresponding value.

**Arguments of `env_condition()`:**

- `MAT` — Mean Annual Temperature in °C. Numeric scalar *or* `function(t)`.
- `MAP` — Mean Annual Precipitation in mm/year. Numeric scalar *or* `function(t)`.

The pattern below simulates a linear 6 °C warming over 60 years by expressing MAT as a function of time:

```{r climate-fn, eval=FALSE}
# Linear warming from 6 °C to 12 °C over 60 years
warming_env <- env_condition(
  MAT = function(t) 6 + (6 / 60) * t,
  MAP = 1200   # static precipitation
)

s    <- stand(data.frame(size_mm = c(150, 200, 350), species_id = "ABIBAL", plot_size = 400))
mod  <- species_model(s)
pars <- parameters(mod, draw = "mean")
ctrl <- control(years = 60, compute_lambda = TRUE, progress = FALSE)

proj_warming <- project(mod, pars, s, warming_env, ctrl)
plot(proj_warming, type = "lambda")
```

`project()` evaluates `MAT(t)` and `MAP(t)` at each year `t`, so any R function works — step changes, sinusoidal cycles, or empirical trajectories loaded from data.

If the time-varying MAT passes outside the fitted climate envelope for a species, a warning is issued once at projection start.
You can silence it by clipping your MAT trajectory to the range shown in `supported_species()`.

# Propagating parameter uncertainty {#sec-uncertainty}

The Bayesian posterior distribution for each demographic model contains 2000 MCMC samples.
To propagate parameter uncertainty into your results, replace `draw = "mean"` with a sample index.

**`parameters()` argument recap:**

- `draw = "mean"` — posterior mean; deterministic.
- `draw = "random"` — random draw from 1–2000; a seed is auto-generated and stored in `pars$seed`.
- `draw = 42L` — specific draw index (integer 1–2000); reproducible without setting a global seed.
- `seed` — integer or `NULL`. Explicit seed for `draw = "random"`. If `NULL`, one is auto-generated.

Retrieve the seed that was used:

```{r seed-example}
pars_r <- parameters(species_model(stand(
  data.frame(size_mm = c(150, 200), species_id = "ABIBAL", plot_size = 400)
)), draw = "random")
pars_r$seed   # auto-generated seed; save this to reproduce the same draw
```

To estimate uncertainty across many draws:

```{r uncertainty, eval=FALSE}
s   <- stand(data.frame(size_mm = c(150, 200, 350), species_id = "ABIBAL", plot_size = 400))
mod <- species_model(s)
env <- env_condition(MAT = 8, MAP = 1200)

# Draw 100 random posterior samples and compute asymptotic lambda for each
lambda_samples <- replicate(100, {
  pars_i <- parameters(mod, draw = "random")
  lam_i  <- lambda(mod, pars_i, s, env)
  lam_i[["ABIBAL"]]
})

hist(lambda_samples, xlab = expression(lambda),
     main = "Posterior distribution of asymptotic lambda — ABIBAL")
abline(v = 1, lty = 2, col = "grey50")
```

Note that `lambda()` is used here rather than `project()` because we only need the asymptotic $\lambda$ at the current stand state — not a full time-series.
`lambda()` arguments:

- `mod`, `pars`, `stand`, `env` — same as `project()`.
- `ctrl` — optional `ipm_control`. Only `bin_width` and `delta_time` are read; if `NULL`, defaults (`bin_width = 1`, `delta_time = 1`) are used.

For computationally intensive analyses (e.g., 1000 draws × multi-species projections), consider parallelising over draws with `parallel::mclapply()` or `future.apply::future_lapply()`.

# Plot-level random effects {#sec-random-effects-deep}

The hierarchical demographic models include plot-level random intercepts for growth, survival, and recruitment.
These represent unmeasured plot characteristics (soil quality, microclimate, disturbance history) that shift vital rates above or below the population mean.

By default all random effects are set to zero (population-mean predictions).
Use `set_random_effects()` to supply observed or estimated plot-level offsets:

**Arguments of `set_random_effects()`:**

- `pars` — an `ipm_parameters` object (from `parameters()`).
- `values` — either:
  - A `numeric(3)` vector `c(growth_offset, survival_offset, recruitment_offset)` applied to all species in `species`.
  - A named list keyed by species ID, where each element is a `numeric(3)` vector — use this to apply different offsets per species.
- `species` — character vector of species IDs. Only used when `values` is a numeric vector. Defaults to `NULL` (apply to all species).

```{r re-single, eval=FALSE}
# Same offsets for all species
pars_re <- set_random_effects(pars, values = c(0.10, -0.05, 0.02))

# Species-specific offsets
pars_re2 <- set_random_effects(
  pars,
  values = list(
    ABIBAL = c( 0.10, -0.05,  0.00),
    ACERUB = c(-0.05,  0.02,  0.10)
  )
)
```

Random effects accumulate across calls — calling `set_random_effects()` twice on the same `pars` object will overwrite the previous values for each named species.

# Controlling memory with `store_every` {#sec-store-every}

For long projections, storing the full size distribution at every timestep can be memory-intensive.
Use `control(store_every = N)` to save only every N-th timestep:

```{r store-every, eval=FALSE}
# 1000-year projection storing only every 10th year (100 snapshots instead of 1000)
ctrl_sparse <- control(
  years          = 1000,
  store_every    = 10,
  compute_lambda = TRUE,
  progress       = TRUE
)
```

The stored timesteps are available in `proj$years` (e.g., `10, 20, 30, ...`).
Pass any value from `proj$years` to `plot(proj, type = "size_dist", timestep = ...)`.

**Full `control()` argument list:**

| Argument | Default | Description |
|---|---|---|
| `years` | `100` | Number of simulation timesteps |
| `delta_time` | `1` | Duration of each timestep in years |
| `store_every` | `1` | Save stand state every N timesteps |
| `bin_width` | `1` | IPM kernel discretization bin width (mm DBH) |
| `compute_lambda` | `FALSE` | Compute asymptotic $\lambda$ at each step (eigendecomposition; slower) |
| `progress` | `TRUE` | Show CLI progress bar |

# Roadmap: what comes next {#sec-roadmap}

The `forestIPM` package is under active development.
Below is the planned feature roadmap, ordered by implementation phase.

## v2 — Coexistence analysis and CI

- **Multi-species coexistence IPM**: a community kernel supporting invasion analysis.
  Given a resident species at equilibrium, run an invader species at low density and measure its invasion growth rate.
  This directly links to the coexistence theory chapters in this book (see @sec-conditional_lambda and @sec-marginal_lambda).
- **GitHub Actions CI**: matrix build (R 4.2, 4.3, 4.4 × Linux, macOS, Windows) and `covr` coverage report on every pull request.
- **`lambda()` vectorisation**: batch computation of $\lambda$ over a grid of climate and competition conditions without re-building model objects.

## v3 — Remote data and ML decoder

- **Parquet parameter backend**: model parameters will be hosted remotely (AWS S3 or equivalent) as a columnar Parquet file, enabling partial reads via Arrow without downloading the full dataset.
  `parameters()` will transparently fetch only the rows needed for the requested species and draw index.
- **ML decoder for new locations**: a learned spatial model that maps latitude/longitude (and optionally plot covariates) to plot-level random effects.
  This will enable projections at any arbitrary location in eastern North America — not just plots with inventory data.
  Planned architecture: a shallow neural network trained on the posterior random effects from the hierarchical model.

## CRAN submission

The package is targeted for CRAN submission after the remote data layer (v3) is stable.
Prerequisites: passing `R CMD check` with zero errors and warnings, vignettes that render without internet dependency, and `covr` coverage ≥ 80 %.

# Cross-references {#sec-xrefs}

- IPM theory and kernel construction: @sec-ipm_description
- Sensitivity analysis: @sec-sens_analysis
- Conditional $\lambda$ (climate × competition surfaces): @sec-conditional_lambda
- Marginal $\lambda$ (integrated across the range): @sec-marginal_lambda
- Extinction risk: @sec-extinction_risk
