---
title: "Deep dive: multi-species IPMs and the forestIPM roadmap"
---

This chapter extends the single-species quick start in @sec-quickstart to cover multi-species IPMs, time-varying climate, and the development roadmap for the `forestIPM` package.

# Multi-species IPMs {#sec-deepdive}

Running a multi-species IPM requires only one change to the quick-start workflow: include multiple `species_id` values in the `stand()` call.
Here we build a two-species stand combining *Abies balsamea* (ABIBAL) and *Acer rubrum* (ACERUB), a common mixedwood assemblage in eastern North America.

```{r multisp-setup}
library(forestIPM)

# Two-species stand on a 400 m² plot
s_multi <- stand(
  data.frame(
    size_mm    = c(150, 200, 180, 350, 280),
    species_id = c("ABIBAL", "ABIBAL", "ACERUB", "ACERUB", "ABIBAL"),
    plot_size  = 400
  )
)
s_multi
```

`species_model()` detects both species automatically and loads their respective fitted models.
The interspecific competition between them is captured internally via the basal area competition terms.

```{r multisp-run, eval=FALSE}
mod_multi  <- species_model(s_multi)
pars_multi <- parameters(mod_multi, draw = "mean")
env_multi  <- env_condition(MAT = 8, MAP = 1200)
ctrl_multi <- control(years = 100, compute_lambda = TRUE, progress = FALSE)

proj_multi <- project(mod_multi, pars_multi, s_multi, env_multi, ctrl_multi)
```

```{r multisp-plot, eval=FALSE}
#| label: fig-multisp
#| fig-cap: "Lambda trajectories for ABIBAL (balsam fir) and ACERUB (red maple) in a mixed stand."
plot(proj_multi, type = "lambda")
```

The plot shows $\lambda$ trajectories for each species.
Species with a competitive advantage under the given climate and density conditions will maintain $\lambda > 1$ longer before density-dependence brings them to equilibrium.

## Exploring species availability

If you are unsure which species IDs are available, use `supported_species()` to browse the full list:

```{r supported}
supported_species()
```

You can filter for species in a particular region or taxonomic group and use the resulting `species_id` values directly in `stand()`.

# Time-varying climate {#sec-time-varying}

A single `env_condition()` object represents a static climate — useful for exploring steady-state behaviour.
To simulate climate change, you iterate over a sequence of conditions, running `project()` for each step and chaining the outputs.

The pattern below shows how to explore the effect of a gradual 6 °C warming over 60 years by stepping through annual `MAT` values:

```{r climate-loop, eval=FALSE}
library(forestIPM)

# Define a 60-year warming trajectory from MAT=6 to MAT=12
mat_sequence <- seq(6, 12, length.out = 60)

# Build baseline stand and model
s    <- stand(data.frame(size_mm = c(150, 200, 350), species_id = "ABIBAL", plot_size = 400))
mod  <- species_model(s)
pars <- parameters(mod, draw = "mean")
ctrl_1yr <- control(years = 1, compute_lambda = TRUE, progress = FALSE)

# Iterate: update climate each year
results <- vector("list", length(mat_sequence))
for (i in seq_along(mat_sequence)) {
  env_i       <- env_condition(MAT = mat_sequence[i], MAP = 1200)
  results[[i]] <- project(mod, pars, s, env_i, ctrl_1yr)
  # Update stand state from previous projection output
  s <- results[[i]]$stand_end
}

# Inspect lambda at each step
lambdas_warming <- sapply(results, function(r) r$summary$lambda)
plot(mat_sequence, lambdas_warming, type = "l",
     xlab = "MAT (°C)", ylab = expression(lambda),
     main = "ABIBAL response to warming")
abline(h = 1, lty = 2, col = "grey50")
```

This approach lets you ask ecological questions such as:
"At what temperature does *Abies balsamea* cross $\lambda = 1$ (transition from growing to declining population)?"

The same pattern applies to MAP gradients or combined MAT + MAP shifts.

# Propagating parameter uncertainty {#sec-uncertainty}

The posterior distribution of each demographic model contains thousands of MCMC samples.
To propagate uncertainty into your projections, replace `draw = "mean"` with a sample index:

```{r uncertainty, eval=FALSE}
# Draw 50 random posterior samples
lambda_samples <- replicate(50, {
  pars_i <- parameters(mod, draw = sample.int(1000L, 1L))
  lam_i  <- lambda(mod, pars_i, s, env, ctrl)
  lam_i[["ABIBAL"]]$asymptotic_lambda
})

hist(lambda_samples, xlab = expression(lambda), main = "Posterior distribution of asymptotic lambda")
abline(v = 1, lty = 2)
```

For computationally intensive analyses (e.g., 1000 draws × multi-species projections), consider parallelising over draws with `parallel::mclapply()` or `future.apply::future_lapply()`.

# Roadmap: what comes next {#sec-roadmap}

The `forestIPM` package is under active development.
Below is the planned feature roadmap, ordered by implementation phase.

## v2 — Coexistence analysis and CI

- **Multi-species coexistence IPM**: a community kernel supporting invasion analysis.
  Given a resident species at equilibrium, run an invader species at low density and measure its invasion growth rate.
  This directly links to the coexistence theory chapters in this book (see @sec-conditional_lambda and @sec-marginal_lambda).
- **GitHub Actions CI**: matrix build (R 4.2, 4.3, 4.4 × Linux, macOS, Windows) and `covr` coverage report on every pull request.
- **`lambda()` vectorisation**: batch computation of $\lambda$ over a grid of climate and competition conditions without re-building model objects.

## v3 — Remote data and ML decoder

- **Parquet parameter backend**: model parameters will be hosted remotely (AWS S3 or equivalent) as a columnar Parquet file, enabling partial reads via Arrow without downloading the full 300 MB dataset.
  `parameters()` will transparently fetch only the rows needed for the requested species and draw.
- **ML decoder for new locations**: a learned spatial model that maps latitude/longitude (and optionally plot covariates) to plot-level random effects.
  This will enable projections at any arbitrary location in eastern North America — not just plots with inventory data.
  Planned architecture: a shallow neural network trained on the posterior random effects from the hierarchical model.

## CRAN submission

The package is targeted for CRAN submission after the remote data layer (v3) is stable.
Prerequisites: passing `R CMD check` with zero errors and warnings, vignettes that render without internet dependency, and `covr` coverage ≥ 80 %.

# Cross-references {#sec-xrefs}

- IPM theory and kernel construction: @sec-ipm_description
- Sensitivity analysis: @sec-sens_analysis
- Conditional $\lambda$ (climate × competition surfaces): @sec-conditional_lambda
- Marginal $\lambda$ (integrated across the range): @sec-marginal_lambda
- Extinction risk: @sec-extinction_risk
