# Gauss-Legendre Quadrature Convergence {#sec-gl_convergence}

```{r,include=FALSE,echo=FALSE}
Echo=FALSE
Eval=TRUE
Cache=TRUE
Warng=FALSE
Msg=FALSE
library(tidyverse)
library(kableExtra)
```

```{r setup,include=FALSE,echo=FALSE}
GL_SIZES <- c(20L, 50L, 100L, 200L, 500L)
GL_REF   <- 500L
GL_TOL_N <- 0.05   # 5% relative tolerance for equilibrium N
```

The release of `v1.0.1` of the `{forestIPM}` R package introduces a new integration approach: the [Gauss-Legendre Quadrature](https://en.wikipedia.org/wiki/Gauss%E2%80%93Legendre_quadrature) — a spectral method that achieves exponential convergence for smooth integrands.
The number of quadrature nodes (`n_gl`) controls the trade-off between accuracy and
computational cost.

Because Gauss-Legendre quadrature exactly integrates polynomials of degree
$\leq 2n-1$, convergence for smooth kernels is typically achieved at modest
$n_{\text{gl}}$ values.
In this chapter, we quantify that convergence empirically using Gauss-Legendre with $n_{gl} = 500$ nodes as the reference.

## Simulation design

Results were pre-computed by the [`compare_gl_N.R`](https://github.com/willvieira/forestIPM/blob/master/simulations/compare_versions/compare_gl_N.R) script:

- **15 stratified environmental scenarios** drawn from 216 027 combinations
  of species × plot × climate.
- **100 Bayesian posterior parameter draws** per scenario.
- **`n_gl` tested:** 20, 50, 100, 200, 500.
- **Reference:** GL(500) (treated as truth for within-GL convergence).
- **Metrics:**
  - Equilibrium $N$ — total tree density after density-dependent projection
    (up to 500 yr, convergence tolerance $10^{-6}$).
  - Equilibrium basal area (BA, m² ha⁻¹).

```{r load_data,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
tbl_N_BA <- readRDS(file.path("data", "table_gl_N_BA.RDS"))

# Reference: GL(500) per (array_id, species)
ref_vals <- tbl_N_BA |>
  filter(method == "GL", n_gl == GL_REF) |>
  select(array_id, species, N_ref = N, BA_ref = BA)

# GL-only table joined with GL(500) reference
gl_tbl <- tbl_N_BA |>
  filter(method == "GL") |>
  left_join(ref_vals, by = c("array_id", "species")) |>
  mutate(
    N_reldev  = abs(N  - N_ref)  / N_ref,
    BA_reldev = abs(BA - BA_ref) / BA_ref,
    sp_short  = sub(".*_", "", species)
  )
```

## Convergence of equilibrium N {#sec-N}

### Relative deviation vs. GL(500)

```{r fig_N_deviation,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| fig-cap: "Relative deviation of equilibrium N from GL(500), by n_gl. Boxes show median and IQR; red dashed = 1%, orange dotted = 5% tolerance."

gl_tbl |>
  filter(n_gl < GL_REF) |>
  mutate(n_gl = factor(n_gl)) |>
  ggplot(aes(n_gl, N_reldev)) +
  geom_boxplot(outlier.size = 0.8, fill = "steelblue", alpha = 0.6) +
  geom_hline(yintercept = 0.01, linetype = "dashed", color = "firebrick",  linewidth = 0.7) +
  geom_hline(yintercept = 0.05, linetype = "dotted", color = "darkorange", linewidth = 0.7) +
  scale_y_log10(
    labels = scales::label_percent(accuracy = 0.01),
    breaks = c(1e-5, 1e-4, 1e-3, 0.01, 0.05, 0.1, 0.5)
  ) +
  labs(
    x       = "n_gl",
    y       = "Relative deviation |N(n) - N(500)| / N(500)",
    caption = "Red dashed = 1%, orange dotted = 5% tolerance"
  ) +
  theme_classic(base_size = 12)
```

### Per-species deviation heatmap {#sec-heatmap}

```{r fig_heatmap_N,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| fig-cap: "Median relative deviation of equilibrium N from GL(500), per species and n_gl."

gl_tbl |>
  filter(n_gl < GL_REF) |>
  group_by(sp_short, n_gl) |>
  summarise(med_dev = median(N_reldev, na.rm = TRUE), .groups = "drop") |>
  mutate(
    n_gl    = factor(n_gl),
    pct_lab = scales::percent(med_dev, accuracy = 0.1)
  ) |>
  ggplot(aes(n_gl, reorder(sp_short, med_dev))) +
  geom_tile(aes(fill = med_dev), color = "white") +
  geom_text(aes(label = pct_lab), size = 2.8) +
  scale_fill_gradient2(
    low      = "#2166ac",
    mid      = "#f7f7f7",
    high     = "#d6604d",
    midpoint = 0.01,
    labels   = scales::label_percent(accuracy = 0.1),
    name     = "Rel. dev."
  ) +
  labs(x = "n_gl", y = NULL) +
  theme_classic(base_size = 11)
```

### Pass/fail at 5% tolerance {#sec-passfail}

```{r tbl_passfail,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
gl_tbl |>
  filter(n_gl < GL_REF) |>
  group_by(n_gl) |>
  summarise(
    n_total    = n(),
    n_pass     = sum(N_reldev < 0.05, na.rm = TRUE),
    pct_pass   = n_pass / n_total,
    max_dev    = max(N_reldev,    na.rm = TRUE),
    median_dev = median(N_reldev, na.rm = TRUE),
    .groups    = "drop"
  ) |>
  mutate(
    pct_pass   = scales::percent(pct_pass,   accuracy = 0.1),
    max_dev    = scales::percent(max_dev,    accuracy = 0.01),
    median_dev = scales::percent(median_dev, accuracy = 0.01)
  ) |>
  rename(
    "n_gl"        = n_gl,
    "N scenarios" = n_total,
    "Pass (< 5%)" = n_pass,
    "% Pass"      = pct_pass,
    "Max dev."    = max_dev,
    "Median dev." = median_dev
  ) |>
  kbl(caption = "Equilibrium N convergence relative to GL(500)") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Convergence of basal area {#sec-BA}

```{r fig_BA_deviation,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| fig-cap: "Relative deviation of equilibrium basal area (BA) from GL(500)."

gl_tbl |>
  filter(n_gl < GL_REF, is.finite(BA_reldev)) |>
  mutate(n_gl = factor(n_gl)) |>
  ggplot(aes(n_gl, BA_reldev)) +
  geom_boxplot(outlier.size = 0.8, fill = "#2ca25f", alpha = 0.6) +
  geom_hline(yintercept = 0.01, linetype = "dashed", color = "firebrick",  linewidth = 0.7) +
  geom_hline(yintercept = 0.05, linetype = "dotted", color = "darkorange", linewidth = 0.7) +
  scale_y_log10(
    labels = scales::label_percent(accuracy = 0.01),
    breaks = c(1e-5, 1e-4, 1e-3, 0.01, 0.05, 0.1, 0.5)
  ) +
  labs(
    x       = "n_gl",
    y       = "Relative deviation |BA(n) - BA(500)| / BA(500)",
    caption = "Red dashed = 1%, orange dotted = 5% tolerance"
  ) +
  theme_classic(base_size = 12)
```

## Per-species convergence profiles {#sec-species}

Each panel shows how both equilibrium N (blue) and BA (green) converge toward
GL(500) as `n_gl` increases.  The ribbon spans the IQR across the 15 scenarios
× 100 parameter draws.

```{r fig_species_facets,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| fig-cap: "Per-species convergence of equilibrium N and BA toward GL(500). Ribbon = IQR across scenarios; red dashed = 1% tolerance."

gl_tbl |>
  filter(n_gl < GL_REF) |>
  pivot_longer(
    cols      = c(N_reldev, BA_reldev),
    names_to  = "metric",
    values_to = "reldev"
  ) |>
  mutate(
    metric = recode(metric, N_reldev = "Equil. N", BA_reldev = "Equil. BA")
  ) |>
  filter(is.finite(reldev)) |>
  group_by(sp_short, n_gl, metric) |>
  summarise(
    med = median(reldev, na.rm = TRUE),
    q25 = quantile(reldev, 0.25, na.rm = TRUE),
    q75 = quantile(reldev, 0.75, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(n_gl, med, color = metric, fill = metric)) +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.2, color = NA) +
  geom_line() +
  geom_point(size = 1.8) +
  geom_hline(yintercept = 0.01, linetype = "dashed", color = "firebrick", linewidth = 0.4) +
  facet_wrap(~sp_short, ncol = 4, scales = "free_y") +
  scale_color_manual(values = c("Equil. N" = "steelblue", "Equil. BA" = "#2ca25f")) +
  scale_fill_manual( values = c("Equil. N" = "steelblue", "Equil. BA" = "#2ca25f")) +
  scale_y_log10(labels = scales::label_percent(accuracy = 0.01)) +
  scale_x_continuous(breaks = c(20, 50, 100, 200)) +
  labs(
    x       = "n_gl",
    y       = "Rel. deviation from GL(500)",
    color   = NULL,
    fill    = NULL,
    caption = "Red dashed = 1% tolerance; ribbon = IQR across scenarios"
  ) +
  theme_classic(base_size = 9) +
  theme(
    strip.text   = element_text(size = 7),
    legend.position = "top"
  )
```

## Species-specific convergence table {#sec-sp-table}

Minimum `n_gl` needed to reach < 1% and < 5% median relative deviation from
GL(500), for both N and BA:

```{r tbl_sp_threshold,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
threshold_tbl <- function(devvar, label) {
  gl_tbl |>
    filter(n_gl < GL_REF) |>
    group_by(sp_short, n_gl) |>
    summarise(med_dev = median(.data[[devvar]], na.rm = TRUE), .groups = "drop") |>
    arrange(sp_short, n_gl) |>
    group_by(sp_short) |>
    summarise(
      min_ngl_1pct = {
        pass <- n_gl[med_dev < 0.01]
        if (length(pass) == 0) NA_integer_ else min(pass)
      },
      min_ngl_5pct = {
        pass <- n_gl[med_dev < 0.05]
        if (length(pass) == 0) NA_integer_ else min(pass)
      },
      dev_100 = scales::percent(med_dev[n_gl == 100], accuracy = 0.01),
      dev_200 = scales::percent(med_dev[n_gl == 200], accuracy = 0.01),
      .groups = "drop"
    ) |>
    rename(
      "Species"                        = sp_short,
      !!paste0("Min n_gl <1% (", label, ")") := min_ngl_1pct,
      !!paste0("Min n_gl <5% (", label, ")") := min_ngl_5pct,
      !!paste0("Dev @100 (", label, ")")      := dev_100,
      !!paste0("Dev @200 (", label, ")")      := dev_200
    )
}

threshold_tbl("N_reldev",  "N") |>
  left_join(threshold_tbl("BA_reldev", "BA"), by = "Species") |>
  kbl(caption = "Minimum n_gl to meet convergence thresholds vs. GL(500)") |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) |>
  add_header_above(c(" " = 1, "Equil. N" = 4, "Equil. BA" = 4))
```

## Recommendation {#sec-recommendation}

```{r recommendation,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| output: asis

pass_summary <- gl_tbl |>
  filter(n_gl < GL_REF) |>
  group_by(n_gl) |>
  summarise(
    pct_N_1pct  = mean(N_reldev  < 0.01, na.rm = TRUE),
    pct_N_5pct  = mean(N_reldev  < 0.05, na.rm = TRUE),
    pct_BA_1pct = mean(BA_reldev < 0.01, na.rm = TRUE),
    pct_BA_5pct = mean(BA_reldev < 0.05, na.rm = TRUE),
    .groups = "drop"
  )

pct <- function(n_gl_val, col) {
  scales::percent(
    pass_summary[[col]][pass_summary$n_gl == n_gl_val],
    accuracy = 0.1
  )
}

cat(glue::glue("
Based on the convergence analysis (15 scenarios × 100 parameter draws,
GL(500) as reference):

| n_gl | N: pass < 1% | N: pass < 5% | BA: pass < 1% | BA: pass < 5% | Recommended use |
|------|-------------|-------------|--------------|--------------|-----------------|
| 20  | {pct(20,  'pct_N_1pct')} | {pct(20,  'pct_N_5pct')} | {pct(20,  'pct_BA_1pct')} | {pct(20,  'pct_BA_5pct')} | **Exploration only** |
| 50  | {pct(50,  'pct_N_1pct')} | {pct(50,  'pct_N_5pct')} | {pct(50,  'pct_BA_1pct')} | {pct(50,  'pct_BA_5pct')} | **Fast screening** |
| 100 | {pct(100, 'pct_N_1pct')} | {pct(100, 'pct_N_5pct')} | {pct(100, 'pct_BA_1pct')} | {pct(100, 'pct_BA_5pct')} | **good balance** |
| 200 | {pct(200, 'pct_N_1pct')} | {pct(200, 'pct_N_5pct')} | {pct(200, 'pct_BA_1pct')} | {pct(200, 'pct_BA_5pct')} | default **Production** |

### Recommended defaults

- **Population projections / equilibrium N and BA:** `n_gl = 200` is the
  recommended default.  It matches GL(500) within 1% for {pct(200, 'pct_N_1pct')}
  of N scenarios and {pct(200, 'pct_BA_1pct')} of BA scenarios.
- **Large-scale simulations (100 k+ runs):** `n_gl = 100` is a viable
  compromise; see species-specific tables for worst-case species.
- **Fast screening / exploration:** `n_gl = 50` is acceptable for
  qualitative comparisons only.

The current package default (`n_gl = 200`) is well-justified and does
not need to be changed.
"))
```

## GL vs. midpoint rule {#sec-vs-midpoint}

Even at GL(500) — which is converged within the GL family — the outputs differ
systematically from the midpoint rule (h = 1 mm).  This section quantifies
that structural difference.

```{r gl_vs_midpoint,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
# Join each GL result with the midpoint reference
midpt_ref <- tbl_N_BA |>
  filter(method == "midpoint") |>
  select(array_id, species, N_midpt = N, BA_midpt = BA)

gl_vs_mp <- tbl_N_BA |>
  filter(method == "GL") |>
  left_join(midpt_ref, by = c("array_id", "species")) |>
  mutate(
    N_dev_mp  = abs(N  - N_midpt)  / N_midpt,
    BA_dev_mp = abs(BA - BA_midpt) / BA_midpt,
    sp_short  = sub(".*_", "", species)
  )
```

### Overall deviation from midpoint

```{r fig_gl_vs_mp_overall,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| fig-cap: "Relative deviation of GL(n) from the midpoint rule (h = 1) for equilibrium N (blue) and BA (green). Unlike within-GL convergence, deviations do not approach zero as n_gl increases — they plateau at a structural difference between the two integration schemes."

gl_vs_mp |>
  filter(is.finite(N_dev_mp), is.finite(BA_dev_mp)) |>
  pivot_longer(
    cols      = c(N_dev_mp, BA_dev_mp),
    names_to  = "metric",
    values_to = "reldev"
  ) |>
  mutate(
    metric = recode(metric, N_dev_mp = "Equil. N", BA_dev_mp = "Equil. BA"),
    n_gl   = factor(n_gl)
  ) |>
  ggplot(aes(n_gl, reldev, fill = metric)) +
  geom_boxplot(outlier.size = 0.6, alpha = 0.6, position = position_dodge(0.8)) +
  scale_fill_manual(values = c("Equil. N" = "steelblue", "Equil. BA" = "#2ca25f")) +
  scale_y_log10(
    labels = scales::label_percent(accuracy = 0.01),
    breaks = c(1e-4, 1e-3, 0.01, 0.05, 0.1, 0.5, 1)
  ) +
  labs(
    x    = "n_gl",
    y    = "Relative deviation from midpoint rule",
    fill = NULL
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")
```

```{r tbl_gl_vs_mp,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
gl_vs_mp |>
  group_by(n_gl) |>
  summarise(
    N_median  = median(N_dev_mp,  na.rm = TRUE),
    N_max     = max(N_dev_mp,     na.rm = TRUE),
    BA_median = median(BA_dev_mp, na.rm = TRUE),
    BA_max    = max(BA_dev_mp,    na.rm = TRUE),
    .groups   = "drop"
  ) |>
  mutate(across(N_median:BA_max, ~scales::percent(.x, accuracy = 0.01))) |>
  rename(
    "n_gl"           = n_gl,
    "N: median dev." = N_median,
    "N: max dev."    = N_max,
    "BA: median dev." = BA_median,
    "BA: max dev."   = BA_max
  ) |>
  kbl(caption = "Deviation of GL(n) from midpoint rule (h = 1)") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Why N deviates more than BA

The two metrics respond differently because of how integration weights enter
each quantity.

**Equilibrium N** is computed as

$$
N^* = \sum_i w_i\, n^*(x_i)
$$

where $w_i$ are quadrature weights and $n^*(x_i)$ is the equilibrium size
density.  For GL nodes, the weights $w_i$ are non-uniform — they are larger
near the centre of the interval and smaller near the endpoints — whereas
midpoint weights are all equal to $h$.  Because small trees are the most
abundant size class and dominate total N, the unequal weighting of GL nodes
in the small-size region directly shifts the integral of the density
distribution, producing a systematic offset in $N^*$ relative to the
uniform midpoint bins.

**Equilibrium BA** is computed as

$$
\text{BA}^* = \sum_i w_i\, n^*(x_i)\, \pi \left(\frac{x_i}{2}\right)^2 \times 10^{-6}
$$

The $x_i^2$ term down-weights small trees and up-weights large trees.  The
large-tree region is where GL weights and midpoint weights are most similar
(the midpoint bins are fine enough to approximate GL there), so the
$x^2$-weighted integral is less sensitive to the quadrature scheme.
As a result, the structural difference between GL and midpoint manifests
more weakly in BA than in N.

In practical terms, **GL and midpoint are computing fundamentally different
discrete approximations** of the same continuous integral.  Neither is wrong;
they converge to the same limit as resolution increases.  However, at finite
resolution, GL distributes nodes and weights optimally for smooth integrands,
while midpoint uses uniform spacing.  For population projections, this means
that **switching from midpoint to GL will always produce a step-change in
equilibrium N and BA outputs**, regardless of how large `n_gl` is.  The
magnitude of this step-change decreases slowly as GL resolution increases,
but does not vanish at practical node counts.


## Computational note

GL quadrature achieves the same within-family accuracy as the midpoint rule
with far fewer nodes.  For example:

- Midpoint (h = 1) for *Quercus rubra* ($L_{\max} \approx 700$ mm) requires
  **573 bins**; GL(100) uses only **100 nodes** — a 5.7× reduction in matrix
  size (32.5× fewer matrix elements).
- Kernel matrix assembly and eigenvalue computation scale as
  $\mathcal{O}(n^2)$ and $\mathcal{O}(n^3)$ respectively, so the savings
  compound quickly with node count.


## Methods summary

| Item | Value |
|------|-------|
| GL algorithm | Golub-Welsch (Jacobi tridiagonal eigendecomposition) |
| C++ solver | `Eigen::SelfAdjointEigenSolver` via RcppEigen |
| Size domain | $[127\,\text{mm},\, L_{\max}]$ (species-specific $L_{\max}$) |
| Vital rate models | von Bertalanffy growth × logistic survival + log-linear recruitment |
| Within-GL reference | GL(500) |
| Midpoint rule | h = 1 mm (original integration method) |
| Scenarios | 15 stratified rows from 216 027 combinations |
| Draws per scenario | 100 Bayesian posterior samples |
| Convergence criterion | $\|\Delta N / N\| < 10^{-6}$ for 10 consecutive steps |
