---
title: "Using the forestIPM package"
---

The `forestIPM` package provides a clean R interface for running Integral Projection Models (IPMs) for 31 tree species in eastern North America.
This guide walks through the complete workflow — from installing the package to projecting a population forward in time and visualizing the results.

# Part 1: Quick start {#sec-quickstart}

## Setup

Install the package from GitHub:

```{r install, eval=FALSE}
# install.packages("devtools")
devtools::install_github("willvieira/forestIPM")
```

```{r load}
library(forestIPM)
```

To see all 31 supported species, call `supported_species()`:

```{r species}
supported_species()
```

Each row lists the species ID (e.g., `ABIBAL` for *Abies balsamea* — balsam fir), the common name, nominal french name, and the observed climate envelope (MAT and MAP min/max) used during model fitting.
The `species_id` column is the key you will use throughout the API.

## Single-species IPM for balsam fir (ABIBAL)

The `forestIPM` workflow is built around five constructors and two engines.
The constructors capture the ecological context (the stand, the model, the parameters, the climate, and the simulation settings) while the engines run the actual computation.

### Step 1 — Define the stand with `stand()`

A stand is a collection of trees described by their size (mm DBH) and species identity.
Here we start a small population of *Abies balsamea* with three trees of different sizes on a 400 m² plot:

```{r stand}
s <- stand(
  data.frame(
    size_mm    = c(150, 200, 350),
    species_id = "ABIBAL",
    plot_size  = 400
  )
)
s
```

`stand()` accepts a single `data` argument — a data.frame or tibble with one row per tree.
The three required columns are flexible in their naming:

| Concept | Accepted column names | Unit |
|---|---|---|
| Tree size | `size_mm`, `size`, or `dbh` | mm DBH |
| Species identity | `species_id`, `species`, or `sp` | character ID |
| Plot area | `plot_size` | m² (same value for all rows) |

The minimum DBH is 127 mm — the lower inventory threshold used to fit the models (FIA protocol).
Trees below this threshold will cause an error.
Use `summary(s)` to inspect the size range and species counts after construction.

### Step 2 — Load the species model with `species_model()`

`species_model()` reads the fitted Bayesian hierarchical demographic model for every species present in the stand:

```{r model}
mod <- species_model(s)
```

**Arguments:**

- `x` — an `ipm_stand` object *or* a character vector of species IDs (e.g., `species_model(c("ABIBAL", "ACERUB"))`).
- `on_missing` — what to do if a species ID is not in `supported_species()`. Defaults to `"error"`. Options:
  - `"error"` (default): abort with a suggestion of the closest valid ID.
  - `"drop"`: silently remove the unrecognised species; the model runs for the remaining species only.
  - `"static"`: keep the species as a competitor whose population does not update over time (useful for fixed background competition).

Use `summary(mod)` to confirm how many species loaded successfully.

### Step 3 — Extract parameters with `parameters()`

`parameters()` returns one parameter realization from the fitted posterior distribution:

```{r parameters}
pars <- parameters(mod, draw = "mean")
```

**Arguments:**

- `mod` — the `ipm_spModel` object from Step 2.
- `draw` — which parameter set to use. Three options:
  - `"mean"` — the posterior mean; deterministic and fastest. Use for exploration and plotting.
  - `"random"` — a random draw from the 2000 posterior samples. A seed is auto-generated and stored in `pars$seed` for reproducibility.
  - Integer `1L` – `2000L` — a specific draw index. Use `set.seed()` before calling, or pass `seed` explicitly.
- `seed` — integer seed for reproducibility when `draw = "random"`. If `NULL` (default), a seed is auto-generated. Retrieve the used seed from `pars$seed`.

```{r parameters-draw}
# Reproducible random draw: fix seed
pars_r <- parameters(mod, draw = "random", seed = 42)

# Specific posterior draw (e.g., draw #137)
pars_137 <- parameters(mod, draw = 137)
```

Plot-level random effects (growth, survival, and recruitment offsets estimated per plot in the hierarchical model) default to zero for all species.
To override them use `set_random_effects()` — see @sec-random-effects.

### Step 4 — Set climate conditions with `env_condition()`

`env_condition()` defines the environmental drivers under which the IPM will run:

```{r env}
env <- env_condition(MAT = 8, MAP = 1200)
```

**Arguments:**

- `MAT` — Mean Annual Temperature in °C. Accepts a numeric scalar (static climate) or a `function(t)` returning a value for year `t` (time-varying climate; see @sec-time-varying).
- `MAP` — Mean Annual Precipitation in mm/year. Same scalar-or-function flexibility.

If your MAT or MAP values fall outside the range observed during model fitting for any species in the stand, `project()` and `lambda()` will emit a warning.
The valid climate envelope for each species is listed in `supported_species()` (columns `MAT_min`, `MAT_max`, `MAP_min`, `MAP_max`).

### Step 5 — Configure the simulation with `control()`

`control()` sets all numerical parameters for the projection engine:

```{r control}
ctrl <- control(years = 200, compute_lambda = TRUE, progress = FALSE)
```

**Arguments (all have defaults):**

| Argument | Default | Description |
|---|---|---|
| `years` | `100` | Number of simulation timesteps to run |
| `delta_time` | `1` | Duration of each timestep in years |
| `store_every` | `1` | Save stand state every N timesteps (reduces memory for long runs) |
| `bin_width` | `1` | Kernel discretization bin width in mm DBH |
| `compute_lambda` | `FALSE` | Compute asymptotic $\lambda$ via eigendecomposition at each step (slower) |
| `progress` | `TRUE` | Show a CLI progress bar during projection |

Setting `store_every = 10` for a 1000-year run reduces the stored `$stand_series` list from 1000 to 100 entries.
The available timesteps are in `proj$years` after the projection.

Use `summary(ctrl)` to confirm how many states will be stored.

### Run the projection with `project()`

`project()` advances the population forward in time, updating the kernel at every timestep to account for the current size distribution (density-dependent dynamics):

```{r project, eval=FALSE}
proj <- project(mod, pars, s, env, ctrl)
```

```{r project-load, echo=FALSE}
# Load precomputed output to avoid slow render
proj  <- readRDS(file.path("data", "out_IPM_guide.RDS"))
```

**Arguments:** `mod`, `pars`, `stand`, `env`, `ctrl` — all five constructors in order.

The returned `ipm_projection` object contains:

| Field | Type | Description |
|---|---|---|
| `$species` | character vector | Focal species IDs |
| `$years` | integer vector | Stored timesteps (determined by `store_every`) |
| `$lambda` | named list of numeric vectors | Per-species $\lambda$ at each stored timestep (only if `compute_lambda = TRUE`) |
| `$stand_series` | list of snapshots | Size distribution per species at each stored timestep |
| `$summary` | tibble | Long-format table: `timestep`, `species_id`, `lambda`, `n_trees` |

For a snapshot of the asymptotic $\lambda$ at a single point in time (without projecting the population forward), use `lambda()` instead — see @sec-lambda-fn.

## Visualizing results with `plot()`

Calling `plot()` on a projection renders diagnostic panels.
Use `type` to select a specific panel:

- `type` — character or `NULL`. Selects which panel to render. Options:
  - `NULL` (default): all four panels in sequence.
  - `"lambda"`: population growth rate $\lambda$ over time.
  - `"pop_size"`: total population size (N) over time.
  - `"size_dist"`: size distribution at a single timestep.
  - `"lambda_vs_n"`: phase portrait of $\lambda$ vs N.
- `timestep` — integer or `NULL`. Only applies to `type = "size_dist"`. The stored year to plot (must be one of `proj$years`). Defaults to the last stored timestep.

### Lambda trajectory {#sec-lambda}

```{r plot-lambda}
#| label: fig-guideLambda
#| fig-cap: "Population growth rate ($\\lambda$) over time. Values above 1 indicate growth; below 1 indicate decline."
plot(proj, type = "lambda")
```

$\lambda$ starts above 1 when the stand is sparse (low intraspecific competition) and converges toward 1 as the population reaches carrying capacity.

### Population size trajectory

```{r plot-popsize}
#| label: fig-guidePopSize
#| fig-cap: "Total population size (N, number of individuals) over time."
plot(proj, type = "pop_size")
```

### Size distribution over time {#sec-sizedist}

```{r plot-sizedist-last}
#| label: fig-guideSizeDist
#| fig-cap: "Size distribution of *Abies balsamea* at year 100. The distribution shifts rightward as the cohort matures."
plot(proj, type = "size_dist")           # default: last stored timestep
```

Use `timestep` to inspect an earlier snapshot:

```{r plot-sizedist-t10}
#| label: fig-guideSizeDist10
#| fig-cap: "Size distribution at year 10 — still dominated by the initial small-tree cohort."
plot(proj, type = "size_dist", timestep = 10)
```

### Lambda vs population size {#sec-lambdavsn}

```{r plot-lambdavsn}
#| label: fig-lambdaVsN
#| fig-cap: "Phase portrait of $\\lambda$ against total population size. Colour gradient indicates time."
plot(proj, type = "lambda_vs_n")
```

This phase portrait shows the classic density-dependent trajectory: $\lambda$ is high when population size is low and declines as intraspecific competition intensifies, eventually stabilising near $\lambda = 1$.

## Computing asymptotic lambda without projecting {#sec-lambda-fn}

For a single-point estimate of $\lambda$ at the current stand state — without projecting the population forward — use `lambda()` directly.
This is much faster than `project()` because it solves one eigenvalue problem rather than iterating over hundreds of timesteps:

```{r lambda-fn}
lam <- lambda(mod, pars, s, env)
lam
```

**Arguments:**

- `mod`, `pars`, `stand`, `env` — same as `project()`.
- `ctrl` — optional `ipm_control` object. Only `bin_width` and `delta_time` are used; if `NULL`, package defaults are applied (`bin_width = 1`, `delta_time = 1`).

The return value is an `ipm_lambda` object: a named numeric vector with one $\lambda$ per focal species.
Use `summary(lam)` to see whether each species is growing ($\lambda > 1$), stable ($\lambda = 1$), or declining ($\lambda < 1$).

## Reproducibility {#sec-reproducibility}

Every `ipm_lambda` and `ipm_projection` object carries a `$conditions` record that captures the exact parameter draw, seed, and climate values used to produce the result.
This makes any analysis self-documenting — you can always recover the exact inputs from the output alone.

```{r reproducibility-inspect}
# The conditions record is embedded in the projection object
proj$conditions$draw_type   # "mean", "random", or "user_defined"
proj$conditions$draw        # integer draw index (NULL for "mean")
proj$conditions$seed        # seed used (NULL for "mean")
proj$conditions$MAT         # climate MAT (numeric or "function(t)")
proj$conditions$MAP         # climate MAP
```

`summary()` prints these alongside the demographic results:

```{r reproducibility-summary}
summary(proj)
```

**Reproducible random draws.**
When `draw = "random"`, a seed is auto-generated and stored in both `pars$seed` and `proj$conditions$seed`.
To reproduce a specific run:

```{r reproducibility-random, eval=FALSE}
# First run — seed auto-generated and saved in the output
pars_r    <- parameters(mod, draw = "random")
proj_run1 <- project(mod, pars_r, s, env, ctrl)
saved_seed <- proj_run1$conditions$seed

# Reproduce the identical run later using the saved seed
pars_r2   <- parameters(mod, draw = "random", seed = saved_seed)
proj_run2 <- project(mod, pars_r2, s, env, ctrl)
# proj_run1$summary and proj_run2$summary are identical
```

**Pinning a specific draw.**
To share a reproducible result without saving a seed, use an integer draw index.
The index is stored in the output and is sufficient to reproduce the run:

```{r reproducibility-pinned, eval=FALSE}
pars_pinned <- parameters(mod, draw = 137L)  # always draw #137 from the posterior
proj_pinned <- project(mod, pars_pinned, s, env, ctrl)
proj_pinned$conditions$draw   # 137 — embedded in the output
```

**`ipm_lambda` conditions** are stored as an attribute on the named numeric vector:

```{r reproducibility-lambda, eval=FALSE}
lam <- lambda(mod, pars_r, s, env)
attr(lam, "conditions")$seed   # same seed as pars_r$seed
attr(lam, "conditions")$MAT    # climate conditions at time of call
```

## Plot-level random effects {#sec-random-effects}

The demographic models are hierarchical: each plot has random intercepts for growth, survival, and recruitment that shift the vital rates above or below the population mean.
By default, `parameters()` sets all random effects to zero (population-mean predictions).

To apply plot-specific random effects — for example if you have inventory data from a known plot — use `set_random_effects()`:

```{r random-effects}
# Apply the same offsets to all species in pars
pars_re <- set_random_effects(
  pars,
  values = c(0.1, -0.05, 0.0)   # growth, survival, recruitment offsets
)

# Or apply species-specific offsets via a named list
pars_re2 <- set_random_effects(
  pars,
  values = list(
    ABIBAL = c(0.1, -0.05, 0.0)
  )
)
```

**Arguments:**

- `pars` — an `ipm_parameters` object.
- `values` — either a `numeric(3)` vector applied to the species listed in `species`, or a named list keyed by species ID where each element is a `numeric(3)` vector.
- `species` — character vector of species IDs. Used only when `values` is a numeric vector. If `NULL` (default), the offsets are applied to all species in `pars`.

The three offsets correspond to the random intercepts in the growth, survival, and recruitment sub-models respectively.

# Part 2: Deep dive (bridge to next chapter) {#sec-deepdive-bridge}

The quick start above covers a single species under static climate conditions.
For more advanced use cases — running multi-species IPMs, simulating climate warming scenarios, propagating posterior uncertainty, and an overview of planned package extensions — see @sec-deepdive in the next chapter.

Briefly, `forestIPM` supports:

- **Multi-species stands**: pass a `data.frame` with multiple `species_id` values to `stand()`; `species_model()` and `project()` handle the rest, including interspecific competition.
- **Time-varying climate**: pass a `function(t)` to `env_condition()` to vary MAT or MAP at each timestep.
- **Posterior uncertainty**: replace `draw = "mean"` with `draw = "random"` or a specific integer index to sample different posterior draws.

For details on all three patterns and the package roadmap, continue to [Deep dive](deep_dive_IPM.qmd).
