# Using the forestIPM package {#sec-ipm_guide}

```{r,include=FALSE,echo=FALSE}
Echo=TRUE
Eval=TRUE
Cache=TRUE
Warng=FALSE
Msg=FALSE
```

The `forestIPM` R package provides a clean R interface for the final Integral Projection Models (IPMs) model and its parameters for the 31 tree species in eastern North America.
This guide walks through the complete workflow — from installing the package, setting conditions, to computing population growth rates and projecting a population forward in time, along with some visualization tools.

## Setup

Install the package from GitHub:

```{r install,echo=Echo,eval=FALSE}
# install.packages("devtools")
devtools::install_github("willvieira/forestIPM")
```

```{r load,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
library(forestIPM)
library(tidyverse)
```

We begin by looking which species have fitted parameters and are available in the package:

```{r species,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
supported_species()
```

Each row lists the species ID (e.g., `ABIBAL` for *Abies balsamea* — balsam fir), the common name, nominal french name, the observed climate envelope (MAT and MAP min/max) used during model fitting, and the fitted demographic model from which the parameters are available.

::: {.callout-note}
The `species_id` column is the key you will use throughout the API.
:::

## Single-species IPM

In the first example, we will construct a simple plot condition with a single species to compute population growth rate ($\lambda$) for balsam fir (`ABIBAL`).
The `forestIPM` workflow is built around five constructors and two engines.
The constructors capture the ecological context (the stand, the model, the parameters, the climate, and the simulation settings) while the engines run the actual computation.
There are two engines, one for computing the punctual $\lambda$ for a given condition, and the other for projecting the dynamics through time.


### Step 1: Define the stand with `stand()`

A stand describes a set of interacting individual trees in a given **local condition**.
Specifically, it defined by a collection of trees with their size (mm DBH) and species identity.
Note that the minimum DBH is 127 mm - the lower inventory threshold used to fit the models (FIA protocol).
We also must define the plot size in which the individuals reside.
The `stand()` function accepts a single data frame object with one row per tree, so it can be easily applied to forest inventory databases.
In this example, we will start a small community from scratch with three individuals per two species in a 400 m² plot:

```{r stand,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
plot_i <- tibble(
  size_mm    = runif(n = 6, 127, 400),
  species_id = c(rep("ABIBAL", 3), rep("ACESAC", 3)),
  plot_size  = 400
)
plot_i
```


```{r stand2,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
std <- stand(plot_i)
std
```

::: {.callout-important}
The IPM is constructed in a way that all species within a plot are interacting with each other. This means that increasing the size of the plot degrades the ecological meaning of the competition.
We are using the 400 m² as it's the size of the FIA protocol.
:::


### Step 2: Load the species model with `species_model()`

`species_model()` reads the fitted Bayesian hierarchical demographic model for one or more species present in the stand.
Here we are interested in computing $\lambda$ for the balsam fir:

```{r model,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
mod <- species_model("ABIBAL")
mod
```


Alternatively, you could directly feed the `stand()` output to the `species_model()` to automatically defining all species **with** available parameters to model.
In the case a stand have species without available parameters, there are three possibilities the user can assign to the `on_missing` argument:

- `"error"` (default): abort with a suggestion of the closest valid ID.
- `"drop"`: silently remove the unrecognised species; the model runs for the remaining species only.
- `"static"`: keep the species as a competitor whose population does not update over time (useful for fixed background competition).


### Step 3: Extract parameters with `parameters()`

The `parameters()` returns one parameter realization from the fitted posterior distribution per `mod` species.
The `draw` argument determines whether the user wants the `"mean"` across all 2000 posterior draws, a sampled `"random"` draw, or a specific draw index (ranging between 1-2000).
When `draw = "random"` a random seed is generated and exported to the object for reproducibility.
But the user can also fixed it with the `seed` argument.

```{r parameters,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
pars <- parameters(mod, draw = "mean")
## Reproducible random draw with fixed seed
# parameters(mod, draw = "random", seed = 42)
## Specific posterior draw (e.g., draw #137)
# parameters(mod, draw = 137)
pars
```

Plot-level random effects (growth, survival, and recruitment offsets estimated per plot in the hierarchical model) default to zero for all species.
To override them use `set_random_effects()` — see @sec-random-effects.


### Step 4: Set climate conditions with `env_condition()`

`env_condition()` defines the environmental drivers under which the IPM will run, namely the mean annual temperature and mean annual precipitation:

```{r env,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
env <- env_condition(MAT = 8, MAP = 1200)
env
```

Further than accepting a numeric scalar (static climate), both arguments also accept a `function(t)` returning a value for year `t` (discussed bellow).

::: {.callout-note}
If your MAT or MAP values fall outside the range observed during model fitting for any species in the stand, the engines will emit a warning.
The valid climate envelope for each species is listed in `supported_species()` (columns `MAT_min`, `MAT_max`, `MAP_min`, `MAP_max`).
:::


### Step 5: Configure the simulation with `control()`

`control()` is the central place to set all configuration parameters for both engines.
The `delta_time` is the duration of each timestep in years. In IPMs, it is usually set to 1.
The other arguments are either for the `project()` engine or the discretization approach (both details discussed later).
For the `lambda()` engine, all parameters can be set to defaut.

```{r control,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
ctrl <- control()
ctrl
```


### Step 6: Compute population growth rate with `lambda()`

Once all constructors are defined, we can approximate the population growth rate ($\lambda$) for a given condition:

```{r lambda_fn,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
lam <- lambda(mod, pars, std, env, ctrl)
lam
```


## Multi-species IPM through time

Different from the `lambda()` engine, `project()` advances the population forward in time, updating the kernel at every timestep to account for the current size distribution (density-dependent dynamics).
This engine also depends on the five constructors.
Here we will the same conditions set to run the single species IPM.

If we plugged all predefined conditions into the `project()` engine, it would return an error stating that there are species in the `std` which do not have its parameters in `pars`.
Indeed, `project()` needs parameters for all species present in the stand, and when there is no parameter available, we have to explicit handle it with the `on_missing` argument in `species_model()`.
To avoid the error, we will first update the `mod` and `pars` objects to also account for the sugar maple along with the balsam fir:

```{r update_mod,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
(mod <- species_model(std))
(pars <- parameters(mod, draw = "random"))
```

As we will project population over time, let's take the opportunity to see how we can set a dynamic environment.
We will simulate a 4°C degree increase for the first 100 year and keep it constant for the remaining years as follows:

```{r set_env_dynamic,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
temp_warm <- function(t) {
  if(t <= 100) {
    return( 8 + (4/100) * t)
  }else{
    return( 8 + 4 )
  }
}

env <- env_condition(MAT = temp_warm, MAP = 1200)
env
```

Before running projection, we can take a look in the available controls for this engine:

```{r set_control2,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
ctrl <- control(
  delta_time = 1,
  years = 300,            # only used in `project()`
  store_every = 1,        # only used in `project()`
  compute_lambda = TRUE,  # only used in `project()`
  progress = FALSE,       # only used in `project()`
  integration_method = "gauss-legendre",
  n_gl = 200L,            # for integration = "gauss-legendre"
  bin_width = 1           # for integration = "midpoint"
)
```

`years` define the total simulation length and `store_every` the steps in which it will save population summaries and lambda computation if `compute_lambda = TRUE`.
The remaining three arguments can be set to default given the Gauss-Legendre Quadrature is both more efficient and mathematically more precise than the midpoint rule to numerically integrate the Kernel.

Now that we have set conditions, we can simulate the community dynamics for the specified time frame:

```{r project,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
proj <- project(mod, pars, std, env, ctrl)
summary(proj)
```

## Visualizing project results with `plot()`

Calling `plot()` on a projection renders diagnostic panels.
Use `type` to select a specific panel:

- `type` — character or `NULL`. Selects which panel to render. Options:
  - `NULL` (default): all four panels in sequence.
  - `"lambda"`: population growth rate $\lambda$ over time.
  - `"pop_size"`: total population size (N) over time.
  - `"size_dist"`: size distribution at a single timestep.
  - `"lambda_vs_n"`: phase portrait of $\lambda$ vs N.
- `timestep` — integer or `NULL`. Only applies to `type = "size_dist"`. The stored year to plot (must be one of `proj$years`). Defaults to the last stored timestep.

### Lambda trajectory {#sec-lambda}

```{r,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-guideLambda
#| fig-cap: "Population growth rate ($\\lambda$) over time. Values above 1 indicate growth; below 1 indicate decline."
plot(proj, type = "lambda")

```

$\lambda$ starts above 1 when the stand is sparse (low intraspecific competition) and converges toward 1 as the population reaches carrying capacity.

### Population size trajectory

```{r,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-guidePopSize
#| fig-cap: "Total population size (N, number of individuals) over time."
plot(proj, type = "pop_size")
```

### Size distribution over time {#sec-sizedist}

```{r,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-guideSizeDist
#| fig-cap: "Size distribution of *Abies balsamea* and *Acer saccharum* at the last time step."
plot(proj, type = "size_dist")           # default: last stored timestep
```

Use `timestep` to inspect an earlier snapshot:

```{r,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-guideSizeDist10
#| fig-cap: "Size distribution at year 10."
plot(proj, type = "size_dist", timestep = 10)
```

### Lambda vs population size {#sec-lambdavsn}

```{r,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-lambdaVsN
#| fig-cap: "Phase portrait of $\\lambda$ against total population size. Colour gradient indicates time."
plot(proj, type = "lambda_vs_n")
```

This phase portrait shows the classic density-dependent trajectory: $\lambda$ is high when population size is low and declines as intraspecific competition intensifies, eventually stabilising near $\lambda = 1$.


## Plot-level random effects {#sec-random-effects}

The demographic models are hierarchical: each plot has random intercepts for growth, survival, and recruitment that shift the vital rates above or below the population mean.
By default, `parameters()` sets all random effects to zero (population-mean predictions).

To apply plot-specific random effects — for example if you have inventory data from a known plot — use `set_random_effects()`:

```{r random_effects,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
# Apply the same offsets to all species in pars
pars_re <- set_random_effects(
  pars,
  values = c(0.1, -0.05, 0.0)   # growth, survival, recruitment offsets
)

# Or apply species-specific offsets via a named list
pars_re2 <- set_random_effects(
  pars,
  values = list(
    ABIBAL = c(0.1, -0.05, 0.0)
  )
)
```

**Arguments:**

- `pars` — an `ipm_parameters` object.
- `values` — either a `numeric(3)` vector applied to the species listed in `species`, or a named list keyed by species ID where each element is a `numeric(3)` vector.
- `species` — character vector of species IDs. Used only when `values` is a numeric vector. If `NULL` (default), the offsets are applied to all species in `pars`.

The three offsets correspond to the random intercepts in the growth, survival, and recruitment sub-models respectively.


## Reproducibility {#sec-reproducibility}

Every `ipm_lambda` and `ipm_projection` object carries a `$conditions` record that captures the exact parameter draw, seed, and climate values used to produce the result.
This makes any analysis self-documenting — you can always recover the exact inputs from the output alone.

```{r reproducibility_inspect,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
# The conditions record is embedded in the projection object
names(proj$conditions)
```

## ROADMAP

The next developpement phase can include

- random effects cloud accecibility
- smooth gaussian model to predict random effects for any location
- new functionality to fit and evaluate Stan model for new speceis
- expand to include dynamic modifiers (immigration, fire/insect disturbance, forest management, etc)
- 