---
title: "Using the forestIPM package"
---

The `forestIPM` package provides a clean R interface for running Integral Projection Models (IPMs) for 31 tree species in eastern North America.
This guide walks through the complete workflow — from installing the package to projecting a population forward in time and visualizing the results.

# Part 1: Quick start {#sec-quickstart}

## Setup

Install the package from GitHub:

```{r install, eval=FALSE}
# install.packages("devtools")
devtools::install_github("willvieira/forest-IPM")
```

```{r load}
library(forestIPM)
```

To see all 31 supported species, call `supported_species()`:

```{r species}
supported_species()
```

Each row lists the species ID (e.g., `ABIBAL` for *Abies balsamea* — balsam fir), the common name, and the taxonomic group.
The species ID is the key you will use throughout the API.

## Single-species IPM for balsam fir (ABIBAL)

The `forestIPM` workflow is built around five constructors and two engines.
The constructors capture the ecological context (the stand, the model, the parameters, the climate, and the simulation settings) while the engines run the actual computation.

**Step 1 — Define the stand.**
A stand is a collection of trees described by their size (mm DBH) and species identity.
Here we start a small population of *Abies balsamea* with three trees of different sizes on a 400 m² plot:

```{r stand}
s <- stand(
  data.frame(
    size_mm    = c(150, 200, 350),
    species_id = "ABIBAL",
    plot_size  = 400
  )
)
s
```

The minimum DBH threshold is 127 mm (the FIA inventory protocol lower limit).

**Step 2 — Load the species model.**
`species_model()` reads the fitted Bayesian hierarchical model for the species present in the stand:

```{r model}
mod <- species_model(s)
```

**Step 3 — Extract parameters.**
`parameters()` returns a parameter set from the posterior distribution.
Use `draw = "mean"` for the posterior mean (deterministic, useful for exploration) or `draw = 1L` for a single posterior sample (for uncertainty propagation):

```{r parameters}
pars <- parameters(mod, draw = "mean")
```

**Step 4 — Set climate conditions.**
`env_condition()` defines the environmental conditions under which the IPM will run.
Here we use a temperate broadleaf climate (MAT = 8 °C, MAP = 1200 mm):

```{r env}
env <- env_condition(MAT = 8, MAP = 1200)
```

**Step 5 — Configure the simulation.**
`control()` sets the number of years to project, whether to compute $\lambda$ at each step, and whether to show a progress bar:

```{r control}
ctrl <- control(years = 100, compute_lambda = TRUE, progress = FALSE)
```

**Run the projection.**
`project()` advances the population forward in time, updating the kernel at every step to account for the current size distribution (density-dependent dynamics):

```{r project, eval=FALSE}
proj <- project(mod, pars, s, env, ctrl)
```

```{r project-load, echo=FALSE}
# Load precomputed output to avoid slow render
out  <- readRDS(file.path("data", "out_IPM_guide.RDS"))
proj <- out$proj
```

The `proj` object is of class `ipm_projection`.
Calling `plot()` on it renders three diagnostic panels at once:

```{r plot-all}
#| label: fig-projAll
#| fig-cap: "Population dynamics of *Abies balsamea* over 100 years under temperate broadleaf conditions. Left: population size over time. Centre: size distribution over time (animated slider). Right: phase portrait of $\\lambda$ vs population size."
plot(proj)
```

## Lambda trajectory {#sec-lambda}

```{r plot-lambda}
#| label: fig-guideLambda
#| fig-cap: "Population growth rate ($\\lambda$) over time. Values above 1 indicate growth; below 1 indicate decline."
plot(proj, type = "lambda")
```

The population growth rate $\lambda$ starts above 1 when the stand is sparse (low competition) and converges toward 1 as the population reaches carrying capacity.

## Size distribution over time {#sec-sizedist}

```{r plot-sizedist}
#| label: fig-guideSizeDist
#| fig-cap: "Size distribution of *Abies balsamea* across the modelled DBH range over 100 years. Use the slider to step through time."
plot(proj, type = "size_dist")
```

The size structure shifts rightward over time as the cohort matures and small trees die or grow into larger size classes.

## Lambda vs population size {#sec-lambdavsn}

```{r plot-lambdavsn}
#| label: fig-lambdaVsN
#| fig-cap: "Phase portrait of population growth rate ($\\lambda$) against total population size over time. Colour gradient indicates time."
plot(proj, type = "lambda_vs_n")
```

This phase portrait shows the classic density-dependent trajectory: $\lambda$ is high when population size is low and declines as intraspecific competition intensifies, eventually stabilising near $\lambda = 1$.

# Part 2: Deep dive (bridge to next chapter) {#sec-deepdive-bridge}

The quick start above covers a single species under static climate conditions.
For more advanced use cases — running multi-species IPMs, simulating climate warming scenarios, and an overview of planned package extensions — see @sec-deepdive in the next chapter.

Briefly, `forestIPM` supports:

- **Multi-species stands**: pass a `data.frame` with multiple `species_id` values to `stand()`; `species_model()` and `project()` handle the rest.
- **Time-varying climate**: loop over `env_condition()` calls with different `MAT`/`MAP` values to simulate climate trajectories.
- **Posterior uncertainty**: replace `draw = "mean"` with `draw = i` (integer) to sample different posterior draws and propagate uncertainty.

For details on all three patterns and the package roadmap (ML decoder for new locations, coexistence IPM, CRAN submission), continue to [Deep dive](deep_dive_IPM.qmd).
