# Marginal effect of covariates {#sec-margin-lambda}

```{r,include=FALSE,echo=FALSE}
Echo=FALSE
Eval=TRUE
Cache=TRUE
Warng=FALSE
Msg=FALSE
library(tidyverse)
library(ggdist)
library(ggtext)
library(ggrepel)
library(ggpubr)
library(sf)
```

```{r re_loadPars,echo=Echo,eval=Eval,cache=Cache,cache.lazy=FALSE,warning=Warng,message=Msg}
data_path <- readLines('_data.path')

# simulation output
out <- readRDS(file.path(data_path, 'simulation_data', 'covariates_perturbation.RDS'))

## Species if info
spIds <- read_csv(
  file.path(
    data_path, 'species_id.csv'
  )
) |>
mutate(
  shade = factor(shade, levels = c('tolerant', 'intermediate', 'intolerant')),
  shade_sylvics = factor(shade_sylvics, levels = c('very-tolerant',
                                            'tolerant',
                                            'intermediate',
                                            'intolerant',
                                            'very-intolerant'
                                            )),
  succession = factor(succession, levels = c('pioneer',
                                              'intermediate',
                                              'climax'))
) |>
filter(sp_to_analyze)

treeData <- readRDS(file.path(data_path, 'treeData.RDS')) |>
	filter(species_id %in% spIds$species_id_old)


# function to determine if a plot is at lower, center, or upper border
# given how much of the tail we consider border (prob arg) [0-0.5]
which_border <- function(temp, prob = 0.1, naRm = TRUE) {
  temp_range = quantile(temp, probs = c(prob, 1 - prob), na.rm = naRm)
  # output vector with 'center' class
  out_pos = rep('Center', length(temp))
  # lower border
  out_pos[temp < temp_range[1]] = 'Cold'
  out_pos[temp > temp_range[2]] = 'Hot'
  
  return(out_pos)
}

# class of range position (cold, center, hot)
treeData |>
  group_by(species_id, plot_id) |>
  # get a single obs per plot to remove abundance effect
  reframe(bio_01_mean = mean(bio_01_mean)) |>
  group_by(species_id) |>
  mutate(
    border_cl = which_border(bio_01_mean, prob = 0.1)
  ) |>
  select(species_id, plot_id, border_cl) |>
  mutate(border_cl = factor(border_cl, levels = c('Cold', 'Center', 'Hot'))) ->
plotBorder_class

# Average temperature range among plots within each border class (for each species)
treeData |>
      left_join(plotBorder_class) |>
      group_by(species_id, border_cl) |>
      reframe(
        range_pos = median(bio_01_mean, na.rm = TRUE)
      ) |>
      # add small noise to range position because if two species have the exact
      # same mean, ggdist will interpret as the a single distribution (bug?!)
      mutate(range_pos = range_pos + rnorm(n(), 0, 0.001)) ->
sp_range_pos


# Average marginal effect for each species
out |>
  group_by(species_id, rep) |>
  reframe(across(contains('par.'), \(x) mean(x, na.rm = TRUE))) ->
AME_sp


out |>
  group_by(species_id, plot_id, year_measured) |>
  mutate(across(contains('par.'), \(x) mean(log(x)))) |>
  slice_head(n = 1) ->
out_summ
```

This chapter assesses the marginal effects or sensitivity of the population growth rate ($\lambda$) to competition and climate conditions.
We aim to understand how this response changes across each species' temperature (or latitude) range distribution.
We define sensitivity as the partial derivative of $\lambda$ with respect to a covariate $X$, which can take the form of conspecific or heterospecific density-dependence competition and temperature or precipitation climate conditions.
In practice, we quantify sensitivity by slightly increasing each covariate value $X_i$ to $X_i'$ and computing the change in $\lambda$ following the right-hand part of @eq-sens.

$$
	\frac{\partial \lambda_i}{\partial X_i} \approx \frac{\Delta \lambda_i}{\Delta X_i} = \frac{|f(X_i') - f(X_i)|}{X_i' - X_i}
$${#eq-sens}

We perform this process for each species across all plot-year observations $i$ to gauge the sensitivity of $\lambda$ that is proportional to the conditions experienced by the species.
We set the perturbation size to a 1% increase in the normalized scale for each covariate.
For instance, a 1% increase translates to a rise of 0.3°C for Mean Annual Temperature (MAT) and 26 mm for Mean Annual Precipitation (MAP).
Because the competition metric is computed at the individual level, the perturbation was applied at each individual, where a 1% increase corresponds to a rise of 1.2 cm in dbh.
As we were interested in the absolute difference, the resulting sensitivity value ranges between 0 and infinity, with lower values indicating a lower sensitivity of $\lambda$ to the specific covariate.

We further computed the log ratio between competition and climate ($CCR$) sensitivities to discern their relative effects as follows:

$$
\begin{align*}
&S_{comp, i} = \frac{\partial \lambda_i}{\partial BA_{cons, i}} + \frac{\partial \lambda_i}{\partial BA_{het, i}}\\
&S_{clim, i} = \frac{\partial \lambda_i}{\partial MAT_{i}} + \frac{\partial \lambda_i}{\partial MAP_{i}}\\
&CCR_i = \text{ln} \frac{S_{comp, i}}{S_{clim, i}}
\end{align*}
$${#eq-ccr}

Here, $S$ represents the total sensitivity to competition or climate.
Negative $CCR$ values indicate higher sensitivity of $\lambda$ to climate, while positive values suggest the opposite.
The code used to perform these analysis are hosted at the [`forest-IPM`](https://github.com/willvieira/forest-IPM/tree/master/simulations/covariates_perturbation) GitHub repository.

## Results

In @fig-ame, we show the sensitivity of each species' population growth rate to conspecific and heterospecific competition, temperature, and precipitation.
Among all species, $\lambda$ is more sensitive to temperature, followed by conspecific and heterospecific competition.

```{r fig-ame,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-ame
#| fig-width: 8.5
#| fig-height: 6
#| fig-cap: "Log sensitivity of species population growth rate to each conspecific competition, heterospecific competition, mean annual temperature, and mean annual precipitation across all plot-year observations. The smaller the values, the lower the sensitivity to a covariate."

AME_sp |>
  left_join(
    spIds,
    by = c('species_id' = 'species_id_old')
  ) |>
  pivot_longer(cols = contains('par.')) |>
  ggplot() +
  aes(log(value), fct_reorder(species_name, value)) +
  aes(color = name) +
  stat_pointinterval() +
  theme_classic() +
  theme(
    axis.text.y = element_text(face = 'italic'),
    legend.position = 'none'
  ) +
  scale_color_manual(
      values = c('#7fc97f', '#ffff99', '#386cb0', '#fdc086'),
      labels = c(
        expression(plain('BA')[conspecific]),
        expression(plain('BA')[heterospecific]),
        'Precipitaiton',
        'Temperature'
      )
    ) +
  labs(
    x = 'ln(Sensitivity)',
    y = NULL,
    color = NULL
  ) ->
p1

AME_sp |>
  group_by(species_id) |>
  reframe(across(contains('par.'), mean)) |>
  pivot_longer(cols = contains('par.')) |>
  mutate(
    name = str_replace(name, 'par.', ''),
    name = case_match(
      name,
      'BA_con' ~ 'BA cons',
      'BA_het' ~ 'BA het',
      'temp' ~ 'Temperature',
      'prec' ~ 'Precipitation'
    )
  ) |>
  ggplot() +
  aes(fct_reorder(name, log(value), .desc = TRUE), log(value)) +
  aes(fill = name) +
  geom_boxplot() +
  theme_classic() +
  labs(
    x = '',
    y = 'ln(Sensitivity)'
  ) +
  scale_fill_manual(values = c('#7fc97f', '#ffff99', '#386cb0', '#fdc086')) +
  theme(legend.position = 'none') ->
p2

ggarrange(p1, p2, ncol = 2)
```

To test how the sensitivity to each covariate changed across the species range, we classified each plot into cold, center, or hot temperature ranges based on MAT.
Cold range plots had MAT below the 10th MAT quantile, while hot range plots had temperatures higher than the 90th MAT quantile.
Everything in between was classified as a center plot.
We describe the sensitivity of $\lambda$ among all species grouped by each of the four covariates in @fig-ame_range.
There were no significant differences in $\lambda$ sensitivity among these ranges within each covariate.

```{r fig-ame_range,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-ame_range
#| fig-width: 8
#| fig-height: 4
#| fig-cap: "Sensitivity of species population growth rate to each covariate grouped by the temperature range position. For each species, plots are classified as cold (MAT < 10th quantile), hot (MAT > 90th quantile), and center (10th < MAT < 90th quantile) given their average MAT across the years."

out_summ |>
  left_join(plotBorder_class) |>
  pivot_longer(cols = contains('par.')) |>
  group_by(species_id, name, border_cl) |>
  reframe(ame_ln = mean(value)) |>
  mutate(
    name = str_replace(name, 'par.', ''),
    name = factor(case_match(
      name,
      'BA_con' ~ 'BA conspecific',
      'BA_het' ~ 'BA heterospecific',
      'temp' ~ 'Temperature',
      'prec' ~ 'Precipitation'
    )),
    name = factor(name, levels = levels(name)[c(4, 1, 2, 3)])
  ) |>
  ggplot() +
  aes(border_cl, ame_ln) +
  aes(fill = border_cl) +
  facet_grid(~name) +
  geom_boxplot() +
  scale_fill_manual(values = c('#91bfdb', '#99d594', '#fc8d59')) +
  theme_classic() +
  labs(
    x = NULL,
    y = 'ln(Sensitivity)',
    fill = NULL
  ) +
  theme(legend.position = 'top')
```

We then merged conspecific with heterospecific competition and temperature with precipitation to assess the relative sensitivity between competition and climate following @eq-ccr.
The predominance of climate over competition was present in all three range positions, although $\lambda$ was more sensitive to competition in the cold range compared to the hot range (@fig-ccr).

```{r fig-ccr,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-ccr
#| fig-width: 7
#| fig-height: 4
#| fig-cap: "Relative sensitivity of species population growth rate to competition and climate grouped by the temperature range position. Negative values indicate higher sensitivity of $\\lambda$ to climate, while positive values indicate higher sensitivity to competition."

out_summ |>
  mutate(
    clim = log(exp(par.temp) + exp(par.prec)),
    comp = log(exp(par.BA_con) + exp(par.BA_het)),
    ccr = comp - clim
  ) |>
  left_join(plotBorder_class) |>
  group_by(species_id, border_cl) |>
  reframe(ccr = mean(ccr)) |>
  ggplot() +
  aes(border_cl, ccr) +
  aes(fill = border_cl) +
  geom_boxplot() +
  scale_fill_manual(values = c('#91bfdb', '#99d594', '#fc8d59')) +
  theme_classic() +
  labs(
    x = NULL,
    y = 'Competition/climate ratio',
    fill = NULL
  )
```

When analyzing the sensitivity at the species level, we observed higher temperature and precipitation sensitivity for species at the cold temperature ranges (@fig-tempVsLat).
This result indicates variability in $\lambda$ sensitivity with respect to their position in the temperature range.

```{r fig-tempVsLat,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-tempVsLat
#| fig-width: 7.5
#| fig-height: 4
#| fig-cap: "Sensitivity of species population growth rate to temperature and precipitation across their temperature range position. We defined each species' temperature range position as the median Mean Annual Temperature across all observed plots."

AME_sp |>
  select(!contains('BA')) |>
  pivot_longer(cols = contains('par.')) |>
  left_join(
    treeData |>
      group_by(species_id, plot_id) |>
      reframe(
        bio_01_mean = mean(bio_01_mean, na.rm = TRUE)
      ) |>
      group_by(species_id) |>
      reframe(
        range_pos = median(bio_01_mean, na.rm = TRUE)
      )
  ) |>
  left_join(
    spIds,
    by = c('species_id' = 'species_id_old')
  ) |>
  group_by(species_name, name, range_pos) |>
  mutate(
    name = case_match(
      name,
      'par.temp' ~ 'Temperature',
      'par.prec' ~ 'Precipitation'
    )
  ) |>
  reframe(
    AME = mean(value),
    range_pos = mean(range_pos)
  ) ->
mean_rangeValue

AME_sp |>
  select(!contains('BA')) |>
  pivot_longer(cols = contains('par.')) |>
  left_join(
    treeData |>
      group_by(species_id, plot_id) |>
      reframe(
        bio_01_mean = mean(bio_01_mean, na.rm = TRUE)
      ) |>
      group_by(species_id) |>
      reframe(
        range_pos = median(bio_01_mean, na.rm = TRUE)
      )
  ) |>
  mutate(
    name = case_match(
      name,
      'par.temp' ~ 'Temperature',
      'par.prec' ~ 'Precipitation'
    )
  ) |>
  ggplot() +
  aes(range_pos, log(value)) +
  facet_wrap(~name) +
  stat_pointinterval(alpha = 0.8) +
  geom_smooth(method = 'lm', color = '#018571') +
  geom_text_repel(
    data = mean_rangeValue,
    aes(x = range_pos, y = log(AME), label = species_name),
    alpha = 0.8,
    size = 2,
    fontface = 'italic'
  ) +
  theme_classic() +
  labs(
    x = 'Mean Annual Temperature (°C)',
    y = 'ln(Sensitivity)'
  )
```

In @fig-compClimVsTemp, we analyze how the sensitivity to competition and climate and their ratio change across the temperature range.
We can observe considerable variability among the species.
Overall, species have lower sensitivity to competition compared to climate, and sensitivity to competition seems to be more stable across the temperature range compared to climate sensitivity.

```{r fig-compClimVsTemp,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-compClimVsTemp
#| fig-width: 9
#| fig-height: 8
#| fig-cap: "Sensitivity of species population growth rate to competition (green) and climate (yellow) along with their ratio across the species temperature range."

# Comp and clim effect across the temperature range 
out_summ |>
  ungroup() |>
  mutate(
    clim = log(exp(par.temp) + exp(par.prec)),
    comp = log(exp(par.BA_con) + exp(par.BA_het)),
    ccr = comp - clim
  ) |>
  select(species_id, temp, clim, comp, ccr) |>
  pivot_longer(cols = c(clim, comp, ccr)) |>
  mutate(
    name = factor(case_match(
      name,
      'comp' ~ 'Competition',
      'clim' ~ 'Climate',
      'ccr' ~ 'Competition/climate ratio'
    )),
    name = factor(name, levels = levels(name)[c(2, 1, 3)])
  ) |>
  left_join(spIds, by = c('species_id' = 'species_id_old')) |>
  ggplot() +
  aes(temp, value) +
  aes(color = name) +
  geom_smooth() +
  facet_wrap(~species_name) +
  theme_classic() +
  scale_color_manual(values = c('#d8b365', '#5ab4ac', 'black')) +
  geom_hline(yintercept = 0, alpha = 0.6) +
  theme(
    strip.text = element_text(face = "italic"),
    strip.background = element_blank(),
    legend.position = 'top'
  ) +
  labs(
    x = 'Mean annual temperature (°C)',
    y = NULL,
    color = NULL
  )
```

Although we could not see much difference in the sensitivity to each covariate across the cold, center, and hot range among all species together (@fig-ame_range), when analyzing each species individually, there are significant changes in the sensitivity ratio between competition and climate across the temperature range (@fig-ccr-sp).

```{r fig-ccr-sp,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-ccr-sp
#| fig-width: 9
#| fig-height: 8
#| fig-cap: "Relative sensitivity of species population growth rate between competition and climate for cold (blue), center (green), and hot (red) temperature ranges."

out_summ |>
  left_join(plotBorder_class) |>
  mutate(par.BA_het = ifelse(is.na(par.BA_het), 0, par.BA_het)) |>
  mutate(
    clim = log(exp(par.temp) + exp(par.prec)),
    comp = log(exp(par.BA_con) + exp(par.BA_het)),
    ccr = comp - clim
  ) |>
  left_join(spIds, by = c('species_id' = 'species_id_old')) |>
  ggplot() +
  aes(border_cl, ccr) +
  aes(fill = border_cl) +
  facet_wrap(~species_name) +
  geom_hline(yintercept = 0, linetype = 2, alpha = 0.4) +
  geom_boxplot() +
  scale_fill_manual(values = c('#91bfdb', '#99d594', '#fc8d59')) +
  theme_classic() +
  theme(
		legend.position = 'top',
		strip.background = element_blank(),
		strip.text = element_text(face = "italic")
	) +
  labs(
    x = NULL,
    y = 'Competion/climate ratio',
    fill = NULL
  ) +
  ylim(-4, 2)
```

Given the sensitivity to competition and climate as functions of the range position changes among the species (@fig-ccr-sp), we extend @fig-ame_range by adding the position of each species on the temperature range to answer the question on how the sensitivity to each covariate changes across the species range.
For that, we calculated the median of MAT for each species for all plots within the cold, center, and hot ranges.
This allowed us to account for the difference in range position between the species to avoid comparing, for instance, the cold range of a boreal species with the cold range of a temperature one.
Overall, for the species and range classes, the sensitivity of $\lambda$ to climate was higher than to the competition covariates (@fig-sensBorder_temp).
Except for sensitivity to climate at the hot range, all the other groups followed the pattern of increase in sensitivity as the species' range position was located towards the colder mean annual temperatures.
At the hot range of the species, the sensitivity to competition decreased towards hotter temperatures, while the climate sensitivity remained constant across the range of MAT.
At the cold range of the species, the sensitivity to both competition and climate increased towards colder temperatures, with a higher increase rate for the climate than competition.
Finally, at the center, the sensitivity to climate and competition increased towards colder temperatures, but there was no clear distinction between the two covariates.

```{r fig-sensBorder_temp,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-sensBorder_temp
#| fig-width: 8.5
#| fig-height: 5.5
#| fig-cap: "Bottom panels describe the sensitivity of species population growth rate to competition (green) and climate (yellow) across the cold, center, and hot temperature ranges. The top panels show the log ratio between competition and climate sensitivities, where negative values mean climate sensitivity is relatively higher than competition. We defined each species' temperature range position as the median Mean Annual Temperature across all observed plots for each cold, center, and hot range class. In the bottom panel, species points are grouped by a Multivariate Normal Density function with 75% probability, while in the top panel, the lines represent the 25, 50, and 75% quantile probabilities."

out |>
  # three species that had the lowest AME (low sensitivity to covariates)
  # filter(!species_id %in% c('NAQUEPRI', '19290QUEALB', '27821NYSSYL')) |>
  mutate(
    clim = log(par.temp + par.prec),
    comp = log(par.BA_con + par.BA_het),
    ccr = comp - clim
  ) |>
  left_join(plotBorder_class) |>
  group_by(species_id, border_cl, rep) |>
  mutate(
    Climate = mean(clim),
    Competition = mean(comp),
    CCR = mean(ccr)
  ) |>
  slice_head(n = 1) |>
  pivot_longer(cols = c(Climate, Competition, CCR)) |>
  left_join(    
    treeData |>
      left_join(plotBorder_class) |>
      group_by(species_id, border_cl) |>
      reframe(
        range_pos = median(bio_01_mean, na.rm = TRUE)
      ) |>
      mutate(range_pos = range_pos + rnorm(n(), 0, 0.001))
  ) ->
plot_dt
  
plot_dt |>
  filter(name == 'CCR') |>
  ggplot() +
  aes(range_pos, value) +
  facet_wrap(~border_cl) +
  geom_point(alpha = 0.4) +
  geom_quantile(quantiles = 0.5, alpha = 0.8, linewidth = 1, color = 'black') +
  geom_quantile(quantiles = c(0.25, 0.75), alpha = 0.6, linewidth = .6, color = 'black') +
  geom_hline(yintercept = 0, alpha = 0.5, linetype = 2) +
  theme_classic() +
  labs(
    # y = expression('ln('~over(S[comp], S[clim])~')'),
    y = 'ln(CCR)',
    x = NULL
  ) +
  theme(
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank(),
    axis.line.x = element_blank()
  ) +
  scale_y_continuous(breaks = c(0, -2, -4)) ->
p1

ggpubr::ggarrange(
  p1,
  plot_dt |>
    filter(name != 'CCR') |>
    ggplot() +
    aes(range_pos, value) +
    aes(fill = name, color = name) +
    facet_wrap(~border_cl) +
    stat_pointinterval(alpha = 0.7) +
    ggdensity::geom_hdr(
      aes(color = NULL),
      probs = .75, alpha = .4,
      method = 'mvnorm'
    ) +
    # geom_smooth(method = 'lm', alpha = .8) +
    scale_color_manual(values = c('#d8b365', '#5ab4ac')) +
    scale_fill_manual(values = c('#d8b365', '#5ab4ac')) +
    theme_classic() +
    labs(
      x = 'Mean annual temperature (°C)',
      y = 'ln(Sensitivity)',
      color = '',
      fill = ''
    ) +
    theme(
      legend.position = 'bottom',
      strip.text = element_blank()
    ) +
    scale_y_continuous(breaks = seq(-3, -9, -2)),
  nrow = 2, heights = c(0.4, 1)
)
```

To quantify the sensitivity difference between competition and climate, we analyze how the $ln(CCR)$ for the cold, center, and hot range changes across the temperature range position of each species (@fig-sensBorder_temp).
Overall, the relative sensitivity of $\lambda$ to climate than competition increases towards the cold and hot ranges.
While at the hot range, the increase in sensitivity to climate was due to the decrease in sensitivity to competition, at the cold range, it was due to a higher increase rate of climate than the competition.
This is why we can observe a higher rate of increase in the relative sensitivity to climate in the hot range compared to the cold range.
When analyzing across the species, the relative sensitivity to competition was higher at the center of the MAT range regardless of the cold or hot range positions.

```{r ccrVsTemp,echo=Echo,eval=FALSE,cache=Cache,warning=Warng,message=Msg}
#| label: fig-ccrVsTemp
#| fig-width: 7
#| fig-height: 4
#| fig-cap: "Relative sensitivity of species population growth rate between competition and climate for cold (blue), center (green), and hot (red) temperature ranges. The summary lines are the 25, 50, and 75% quantiles."

out |>
  # three species that had the lowest AME (low sensitivity to covariates)
  # filter(!species_id %in% c('NAQUEPRI', '19290QUEALB', '27821NYSSYL')) |>
  mutate(
    clim = log(par.temp + par.prec),
    comp = log(par.BA_con + par.BA_het),
    ccr = comp - clim
  ) |>
  left_join(plotBorder_class) |>
  group_by(species_id, border_cl, rep) |>
  mutate(ccr = mean(ccr)) |>
  slice_head(n = 1) |>
  left_join(sp_range_pos) |>
  filter(border_cl != 'Center') |>
  left_join(
    spIds,
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(range_pos, ccr) +
  aes(color = border_cl) +
  stat_pointinterval(alpha = 0.7) +
  geom_quantile(quantiles = 0.5, linewidth = 1.2) +
  geom_quantile(quantiles = c(0.25, 0.75), alpha = 0.8, linewidth = .8) +
  geom_hline(yintercept = 0, alpha = 0.5, linetype = 2) +
  scale_color_manual(values = c('#91bfdb', '#fc8d59')) +
  theme_classic() +
  labs(
    x = 'Mean annual temperature (°C)',
    y = 'Competition/climate ratio',
    color = ''
  )
```

## Spatial Distribution of sensitivity to the covariates

So far, we have discussed the sensitivity of $\lambda$ to competition and climate using averages across species and temperature range positions.
Below, we summarize the raw observed sensitivity to competition, climate, and their ratio for each species-plot observation.
For plots with more than one observation, we averaged each of the three metrics across the observations.

```{r funcPrint,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
plot_spatial_sens <- function(sp) {
  out_summ |>
    filter(species_id == sp) |>
    mutate(
      clim_ln = log(exp(par.temp) + exp(par.prec)),
      comp_ln = log(exp(par.BA_con) + exp(par.BA_het)),
      ccr_ln = comp_ln - clim_ln
    ) |>
    group_by(plot_id) |>
    mutate(
      clim_ln = mean(clim_ln, na.rm = TRUE),
      comp_ln = mean(comp_ln, na.rm = TRUE),
      ccr_ln = mean(ccr_ln, na.rm = TRUE)
    ) |>
    slice_head(n = 1) |>
    select(contains('tude'), contains('ln')) |>
    ungroup() ->
  sp_dt

  theme_plot <- list(
    theme_classic() +
    theme(
      legend.position = 'bottom',
      axis.text.y = element_blank(),
      axis.text.x = element_blank(),
			legend.key.height = unit(2, 'mm')
    )
  )

  sp_dt |>
    filter(
      comp_ln > quantile(comp_ln, 0.05) & comp_ln < quantile(comp_ln, 0.95)
    ) |>
    filter(
      clim_ln > quantile(clim_ln, 0.05) & clim_ln < quantile(clim_ln, 0.95)
    ) |>
    ungroup() |>
    select(comp_ln, clim_ln) |>
    pivot_longer(cols = everything()) |>
    pull(value) |>
    range() ->
  sens_range

  sp_dt |>
    filter(
      comp_ln > quantile(comp_ln, 0.05) & comp_ln < quantile(comp_ln, 0.95)
    ) |>
    st_as_sf(
      coords = c('longitude', 'latitude'),
      crs = 4326
    ) |>
    ggplot() +
    geom_sf(aes(color = comp_ln), alpha = 0.7, size = 0.1) +
    scale_color_viridis_c(limits = sens_range) +
    labs(color = expression('ln('~S[competition]~')')) +
    theme_plot ->
  p1

  sp_dt |>
    filter(
      clim_ln > quantile(clim_ln, 0.05) & clim_ln < quantile(clim_ln, 0.95)
    ) |>
    st_as_sf(
      coords = c('longitude', 'latitude'),
      crs = 4326
    ) |>
    ggplot() +
    geom_sf(aes(color = clim_ln), alpha = 0.7, size = 0.1) +
    scale_color_viridis_c(limits = sens_range) +
    labs(color = expression('ln('~S[climate]~')')) +
    theme_plot ->
  p2


  sp_dt |>
    filter(
      ccr_ln > quantile(ccr_ln, 0.05) & ccr_ln < quantile(ccr_ln, 0.95)
    ) |>
    st_as_sf(
      coords = c('longitude', 'latitude'),
      crs = 4326
    ) |>
    ggplot() +
    geom_sf(aes(color = ccr_ln), alpha = 0.7, size = 0.1) +
    scale_color_gradient2() +
    labs(color = expression('ln(CCR)')) +
    theme_plot ->
  p3

  print(
      annotate_figure(
        ggarrange(p1, p2, p3, nrow = 1),
        top = text_grob(
          spIds |> filter(species_id_old == sp) |> pull(species_name),
          face = 'italic', size = 10
        )
      )
  )
}
```


```{r print_plots,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| fig-width: 8.5
#| fig-height: 3.5

for(sp in spIds$species_id_old)
  plot_spatial_sens(sp)
```
