# Metapopulation dynamics {#sec-metapop}

```{r,include=FALSE,echo=FALSE}
Echo=FALSE
Eval=TRUE
Cache=TRUE
Warng=FALSE
Msg=FALSE
library(tidyverse)
library(ggdist)
library(ggpubr)
library(ggrepel)
library(ggdensity)
```

Here we will explore the a simple version of the metapopulation theory and how the colonization and extinction parameters can determine range limits.
Explain the metapop theory here [@Levins1969].

$$
\frac{dp}{dt} = \alpha p (1 - p) - \varepsilon p
$$

We will start with a gradient of mean annual temperature (MAT) ranging from 0 to 20°C, where species are more abundant in colder rathar than hot temperature.
Thefore, colonization  ($\mathcal{C}$) reduces and extinction ($\epsilon$) increases as temperature rises following the simple linear model:

$$
\begin{split}
&\mathcal{C}(MAT) = \beta_{0, \mathcal{C}} + MAT * \beta_{1, \mathcal{C}} \\[2pt]
&\epsilon(MAT) = \beta_{0, \epsilon} + MAT * \beta_{1, \epsilon}
\end{split}
$$ {#eq-parsMAT}

For instance, we assign colonization and extinction probabilities of the focal species to 0.2 and 0.08 ($\beta_0$).
Also, we define that colonization and extinction changes with MAT at a rate of -0.12 and 0.15 ($\beta_1$), therefore decreasing and increasing with MAT, respectively.

```{r simulate_metapop,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
# parameters (C and E respectively)
beta_0 <- c(0.4, 0.1)
beta_1 <- c(-0.02, 0.005)

metapop <- function(p, MAT, sp = 1, beta_0, beta_1) {
  col = beta_0[1] + beta_1[1] * MAT
  ext = beta_0[2] + beta_1[2] * MAT

  col * p * (sp - p) - ext * p
}

tibble(
  beta_0,
  beta_1,
  par = c('Colonization', 'Extinction')
) |>
group_by(par) |>
expand_grid(MAT = seq(0, 20, 0.1)) |>
mutate(
  value = beta_0 + beta_1 * MAT
) |>
ggplot() +
aes(MAT, value) +
aes(color = par) +
geom_line() +
theme_classic() +
theme(legend.position = 'top') +
labs(
  x = 'Mean annual temperature (°C)',
  y = 'Colonization/Extinction',
  color = NULL
)

tibble(
  MAT = seq(0, 20, 0.1)
) |>
group_by(MAT) |>
expand_grid(
  sp = c(0.1, 0.5, 1),
  p = c(0.1, 0.5, 1)
) |>
mutate(
  dpdt = metapop(p, MAT, sp, beta_0, beta_1)
) |>
ggplot() +
aes(MAT, dpdt) +
aes(color = factor(sp)) +
facet_wrap(~p) +
geom_line() +
geom_hline(yintercept = 0, alpha = 0.7) +
theme_classic() +
theme(legend.position = 'top')



expand_grid(
  c = seq(0, 1, 0.01),
  e = seq(0, 1, 0.01)
) |>
group_by(c, e) |>
expand_grid(
  p = 0.01,
  sp = c(0.2, 0.5, 1)
) |>
mutate(K = sp - e/c) |>
ggplot() +
aes(e, c, fill = exp(K)) +
facet_wrap(~sp) +
geom_tile() +
scale_fill_gradient2()

expand_grid(
  sp = seq(0, 1, 0.01)
) |>
group_by(sp) |>
expand_grid(
  c = c(0.2, 0.4, 0.6, 0.8),
  e = c(0.1, 0.2, 0.4)
) |>
# 
mutate(K = exp(sp - e/c)) |>
ggplot() +
aes(sp, K, color = factor(c)) +
facet_wrap(~e) +
geom_line()
```