# Competition effect {#sec-parsComp}

In this section, we discuss the effect of competition on the growth, survival, and recruitment vital rates.
For all three demographic functions, competition for light increases linearly with the plot density, determined by the $\beta$ parameter, indicating the density-dependent effect.
Assuming that light represents the most limiting resource for trees, we define that the influence of individuals smaller than the focal tree is negligible. 
Consequently, we used the total basal area of individuals larger than the focal tree to measure competition.
Furthermore, we divided the impact of the total basal area into contributions from conspecific and heterospecific individuals larger than the focal individual.

```{r,include=FALSE,echo=FALSE}
Echo=FALSE
Eval=TRUE
Cache=TRUE
Warng=FALSE
Msg=FALSE
library(tidyverse)
library(cmdstanr)
library(posterior)
library(ggdist)
library(ggpubr)
library(ggrepel)
```

```{r comp_loadPars,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
data_path <- readLines('_data.path')
pars_path <- file.path(data_path, 'output_sim_processed')
models <- c(
  'growth' = 'intcpt_plot_comp_clim',
  'mort' = 'intcpt_plot_comp_clim',
  'recruit' = 'intcpt_plot_comp_clim'
)

## Species if info
spIds <- read_csv(
  file.path(
    data_path, 'species_id.csv'
  )
) |>
mutate(
  shade = factor(shade, levels = c('tolerant', 'intermediate', 'intolerant')),
  shade_sylvics = factor(shade_sylvics, levels = c('very-tolerant',
                                            'tolerant',
                                            'intermediate',
                                            'intolerant',
                                            'very-intolerant'
                                            )),
  succession = factor(succession, levels = c('pioneer',
                                              'intermediate',
                                              'climax'))
) |>
filter(sp_to_analyze)


# Load parameters
map_dfr(
  spIds$species_id_old,
  ~ readRDS(
    paste0(
      pars_path, '/growth/', models['growth'], '/posterior_pop_', .x, '.RDS'
    )
  ) |>
  filter(par %in% c('Beta', 'theta')) |>
  pivot_wider(names_from = par) |>
  bind_cols(species_id = .x)
) ->
pars_growth

map_dfr(
  spIds$species_id_old,
  ~ readRDS(
    paste0(
      pars_path, '/mort/', models['mort'], '/posterior_pop_', .x, '.RDS'
    )
  ) |>
  filter(par %in% c('Beta', 'theta')) |>
  pivot_wider(names_from = par) |>
  bind_cols(species_id = .x)
) ->
pars_survival

map_dfr(
  spIds$species_id_old,
  ~ readRDS(
    paste0(
      pars_path, '/recruit/', models['recruit'], '/posterior_pop_', .x, '.RDS'
    )
  ) |>
  filter(par %in% c('beta_p', 'optimal_BA', 'sigma_BA')) |>
  pivot_wider(names_from = par) |>
  bind_cols(species_id = .x)
) ->
pars_rec

treeData <- readRDS(file.path(data_path, 'treeData.RDS')) |>
  filter(species_id %in% spIds$species_id_old)
```


## Conspecific and heterospecific effect

#### Growth

```{r comp_compGrowth,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-compGrowth
#| fig-width: 8.5
#| fig-height: 6
#| fig-cap: "Posterior distribution of conspecific and heterospecific density dependence across species (left panel) and species shade tolerance (right panel). Species are classified by their general growth trait following @burns1990silvics"

pars_growth |>
  mutate(
    Con = Beta,
    Het = Beta * theta
  ) |>
  pivot_longer(cols = c(Con, Het)) |>
  left_join(
    spIds,
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(value, fct_reorder(species_name, value)) +
  aes(color = name) +
  stat_pointinterval() +
  geom_vline(xintercept = 0, alpha = 0.7) +
  ylab('') +
  xlab(expression('Density dependence effect ('~beta~')')) +
  scale_color_manual(
    values = c('#ef8a62', '#67a9cf')
  ) +
  theme_classic() +
  theme(
    axis.text.y = element_text(face = "italic"),
    legend.position = 'none'
  ) +
  labs(color = '') ->
p1

pars_growth |>
  group_by(species_id) |>
  reframe(
    Beta = mean(Beta),
    theta = mean(theta)
  ) |>
  mutate(
    Con = Beta,
    Het = Beta * theta
  ) |>
  pivot_longer(cols = c(Con, Het)) |>
  left_join(
    spIds,
    by = c('species_id' = 'species_id_old')
  ) |>
  mutate(
    name = case_match(
      name,
      'Con' ~ 'Conspecific',
      'Het' ~ 'Heterospecific'
    )
  ) |>
  ggplot() +
  aes(shade_sylvics, value) +
  aes(fill = name) +
  geom_boxplot() +
  scale_fill_manual(
    values = c('#ef8a62', '#67a9cf')
  ) +
  xlab('') +
  ylab(expression('Density dependence effect ('~beta~')')) +
  theme_classic() +
  theme(legend.position = 'top') +
  labs(fill = '') ->
p2

ggarrange(p1, p2, ncol = 2)
```


#### Survival

```{r comp_compSurv,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-compSurvival
#| fig-width: 8.5
#| fig-height: 6
#| fig-cap: "Posterior distribution of conspecific and heterospecific density dependence across species (left panel) and species shade tolerance (right panel). Species are classified by their general growth trait following @burns1990silvics"

pars_survival |>
  mutate(
    Con = Beta,
    Het = Beta * theta
  ) |>
  pivot_longer(cols = c(Con, Het)) |>
  left_join(
    spIds,
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(value, fct_reorder(species_name, value)) +
  aes(color = name) +
  stat_pointinterval() +
  geom_vline(xintercept = 0, alpha = 0.7) +
  ylab('') +
  xlab(expression('Density dependence effect ('~beta~')')) +
  scale_color_manual(
    values = c('#ef8a62', '#67a9cf')
  ) +
  theme_classic() +
  theme(
    axis.text.y = element_text(face = "italic"),
    legend.position = 'none'
  ) +
  labs(color = '') ->
p1


pars_survival |>
  group_by(species_id) |>
  reframe(
    Beta = mean(Beta),
    theta = mean(theta)
  ) |>
  mutate(
    Con = Beta,
    Het = Beta * theta
  ) |>
  pivot_longer(cols = c(Con, Het)) |>
  left_join(
    spIds,
    by = c('species_id' = 'species_id_old')
  ) |>
  mutate(
    name = case_match(
      name,
      'Con' ~ 'Conspecific',
      'Het' ~ 'Heterospecific'
    )
  ) |>
  ggplot() +
  aes(shade_sylvics, value) +
  aes(fill = name) +
  geom_boxplot() +
  scale_fill_manual(
    values = c('#ef8a62', '#67a9cf')
  ) +
  xlab('') +
  ylab(expression('Density dependence effect ('~beta~')')) +
  theme_classic() +
  theme(legend.position = 'top') +
  labs(fill = '') ->
p2

ggarrange(p1, p2, ncol = 2)
```


#### Growth vs survival

```{r comp_compGrowthVsSurv,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-compGrowthVsSurv
#| fig-width: 8.5
#| fig-height: 5
#| fig-cap: "Correlation between growth and survival density dependence parameters."

pars_growth |>
  mutate(
    Con_growth = Beta,
    Het_growth = Beta * theta
  ) |>
  select(!c(Beta, theta)) |>
  group_by(species_id) |>
  mutate(iter = 1:n()) |>
  left_join(
    pars_survival |>
      mutate(
        Con_mort = Beta,
        Het_mort = Beta * theta
      ) |>
      select(!c(Beta, theta)) |>
      group_by(species_id) |>
      mutate(iter = 1:n())
  ) |>
  pivot_longer(
    cols = c(contains('Het_'), contains('Con_')),
    names_to = c('comp', 'vr'),
    names_sep = '_'
  ) |>
  mutate(
    comp = case_match(
      comp,
      'Con' ~ 'Conspecific',
      'Het' ~ 'Heterospecific'
    )
  ) |>
  pivot_wider(
    names_from = vr,
    values_from = value
  ) |>
  left_join(
    spIds,
    by = c('species_id' = 'species_id_old')
  ) |>
  group_by(species_id, comp) |>
  slice_sample(n = 1000) |>
  ggplot() +
  aes(growth, mort) +
  # aes(color = shade_sylvics) +
  facet_wrap(~comp) +
  geom_point(alpha = 0.1, size = 0.1) +
  geom_abline(intercept = 0, slope = 1, alpha = 0.5) +
  xlab(expression('Growth density dependence effect ('~beta~')')) +
  ylab(expression('Survival density dependence effect ('~beta~')')) +
  scale_color_manual(
    # values = c("#87bc45", "#edbf33", "#ea5545")
    values = c("#20bc45", "#87bc45", "#edbf33", "#ea5545", "#ba0000")
  ) +
  theme_classic() +
  labs(color = '') +
  tune::coord_obs_pred()
```


## Total basal area effect on recruitment survival

```{r comp_compGrowthVsSurv,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-compRec
#| fig-width: 8.5
#| fig-height: 6
#| fig-cap: "Posterior distribution of the density dependence parameter affecting the annual survival rate of recruitment individuals."

pars_rec |>
  left_join(
    spIds,
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(-beta_p, fct_reorder(species_name, -beta_p)) +
  # aes(color = ) +
  stat_pointinterval() +
  geom_vline(xintercept = 0, alpha = 0.7) +
  ylab('') +
  xlab(expression('Density dependence effect ('~beta~')')) +
  # scale_color_manual(
  #   values = c("#87bc45", "#edbf33", "#ea5545")
  # ) +
  theme_classic() +
  theme(
    axis.text.y = element_text(face = "italic"),
    legend.position = 'none'
  ) +
  labs(color = '') ->
p1


pars_rec |>
  group_by(species_id) |>
  reframe(
    beta_p = mean(beta_p)
  ) |>
  left_join(
    spIds,
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(shade_sylvics, -beta_p) +
  aes(fill = shade_sylvics) +
  geom_boxplot() +
  scale_fill_manual(
    # values = c("#87bc45", "#edbf33", "#ea5545")
    values = c("#20bc45", "#87bc45", "#edbf33", "#ea5545", "#ba0000")
  ) +
  xlab('') +
  ylab(expression('Density dependence effect ('~beta~')')) +
  theme_classic() +
  theme(legend.position = 'none') +
  labs(fill = '') ->
p2

ggarrange(p1, p2, ncol = 2)
```


## How does CNDD relate to species latitude and abundance patterns?

This is mostly inspired by @hulsmann2024, where they test many hypotheses around conspecific density dependence (CNDD) around the world.
Their main finding is that CNDD is stronger in the tropics compared to temperature species, but only for low or intermediate abundant species.

```{r CNDDvsLat,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-CNDDvsLat
#| fig-width: 8.5
#| fig-height: 5
#| fig-cap: "Posterior distribution of the density dependence parameter affecting the annual survival rate of recruitment individuals."

pars_growth |>
  mutate(
    Con_Growth = Beta,
    Het_Growth = Beta * theta
  ) |>
  select(!c(Beta, theta)) |>
  group_by(species_id) |>
  mutate(iter = 1:n()) |>
  left_join(
    pars_survival |>
      mutate(
        Con_Survival = Beta,
        Het_Survival = Beta * theta
      ) |>
      select(!c(Beta, theta)) |>
      group_by(species_id) |>
      mutate(iter = 1:n())
  ) |>
  left_join(
    pars_rec |>
      mutate(
        Con_Recruit = -beta_p,
        Het_Recruit = NA
      ) |>
      select(!c(beta_p, optimal_BA, sigma_BA)) |>
      group_by(species_id) |>
      mutate(iter = 1:n())
  ) |>
  pivot_longer(
    cols = c(contains('Het_'), contains('Con_')),
    names_to = c('comp', 'vr'),
    names_sep = '_'
  ) |>
  mutate(
    comp = case_match(
      comp,
      'Con' ~ 'Conspecific',
      'Het' ~ 'Heterospecific'
    ),
    vr = factor(vr, levels = c('Growth', 'Survival', 'Recruit'))
  ) |>
  left_join(
    treeData |>
      group_by(species_id) |>
      reframe(lat_mean = median(latitude)) 
  ) |>
  ggplot() +
  aes(lat_mean, value) +
  aes(color = comp, fill = comp) +
  facet_wrap(~vr) +
  stat_pointinterval() +
  ggdensity::geom_hdr(
    aes(color = NULL),
    probs = .75, alpha = .4,
    method = 'mvnorm'
  ) +
  scale_color_manual(values = c('#ef8a62', '#67a9cf')) +
  scale_fill_manual(values = c('#ef8a62', '#67a9cf')) +
  geom_hline(yintercept = 0, linetype = 2) +
  theme_classic() +
  labs(
    x = 'Median latitude',
    y = 'Density dependence',
    color = NULL,
    fill = NULL
  ) +
  theme(legend.position = 'top')
```


```{r CNDDvsAbund,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-CNDDvsAbund
#| fig-width: 8.5
#| fig-height: 5
#| fig-cap: "Posterior distribution of the density dependence parameter affecting the annual survival rate of recruitment individuals."

pars_growth |>
  mutate(
    Con_Growth = Beta,
    Het_Growth = Beta * theta
  ) |>
  select(!c(Beta, theta)) |>
  group_by(species_id) |>
  mutate(iter = 1:n()) |>
  left_join(
    pars_survival |>
      mutate(
        Con_Survival = Beta,
        Het_Survival = Beta * theta
      ) |>
      select(!c(Beta, theta)) |>
      group_by(species_id) |>
      mutate(iter = 1:n())
  ) |>
  left_join(
    pars_rec |>
      mutate(
        Con_Recruit = -beta_p,
        Het_Recruit = NA
      ) |>
      select(!c(beta_p, optimal_BA, sigma_BA)) |>
      group_by(species_id) |>
      mutate(iter = 1:n())
  ) |>
  pivot_longer(
    cols = c(contains('Het_'), contains('Con_')),
    names_to = c('comp', 'vr'),
    names_sep = '_'
  ) |>
  mutate(
    comp = case_match(
      comp,
      'Con' ~ 'Conspecific',
      'Het' ~ 'Heterospecific'
    ),
    vr = factor(vr, levels = c('Growth', 'Survival', 'Recruit'))
  ) |>
  left_join(
    treeData |>
      group_by(species_id) |>
      reframe(BA_plot_mean = median(BA_plot, na.rm = T)) 
  ) |>
  ggplot() +
  aes(BA_plot_mean, value) +
  aes(color = comp, fill = comp) +
  facet_wrap(~vr) +
  stat_pointinterval() +
  ggdensity::geom_hdr(
    aes(color = NULL),
    probs = .75, alpha = .4,
    method = 'mvnorm'
  ) +
  scale_color_manual(values = c('#ef8a62', '#67a9cf')) +
  scale_fill_manual(values = c('#ef8a62', '#67a9cf')) +
  geom_hline(yintercept = 0, linetype = 2) +
  theme_classic() +
  labs(
    x = 'Median BA plot',
    y = 'Density dependence',
    color = NULL,
    fill = NULL
  ) +
  theme(legend.position = 'top')
```