# Mixed covariates {#sec-mixed_cov}

```{r,include=FALSE,echo=FALSE}
Echo=FALSE
Eval=TRUE
Cache=TRUE
Warng=FALSE
Msg=FALSE
library(tidyverse)
library(ggtext)
library(ggpubr)
library(MetBrewer)
```

In this study, our main objective is to assess how climate and competition affect the demographic rates of tree species, and hence shaping their range distribution.
Climate, typically characterized by temperature and precipitation, is widely assumed as an important factor affecting vital rates and has been the focus of recent studies [@Csergo2017; @lesquin2021; @Kunstler2021; @Guyennon2023].
Climate also exerts an indirect influence by shaping species composition, which in turn, impacts the variation in demographic rates through species interactions.
indeed, competition for light has been shown as a principal driver in the demographic rates of forest trees [@Zhang2015; @lesquin2021].

There are additional factors that influence the growth, survival, and recruitment of forest trees beyond the climate-competition dimensions.
For instance, events like wildfire and insect outbreaks play crucial roles in changing demographic rates, particularly considering that these disturbances are sensitivite to climate change [@seidl2011].
However, it's important to note that such disturbances are sporadic events, and our primary focus is on understanding responses to average conditions.
At the local scale, soil nitrogen can improve growth rate [@Ibanez2018] and facilitation can improve performance at range limits [@Ettinger2017].
At a local scale, soil nitrogen content can enhance growth rates [@Ibanez2018], and facilitation can increase recruitment rates at range limits [@Ettinger2017].
All these factors and other not cited here have the potential to affect the demographic rates of forest trees.
However, our objective here is not to have the best and most complex model to achive the highest predicitive metric, but to do inference [@Tredennick2021].
Specifically, we aim to test the relative effect of climate and competition while controlling for other influential factors.
Therefore, our modeling approach is guided by biological mechanisms, which tend to provide more robust extrapolation [@Briscoe2019], rather than being solely dictated by specific statistical metrics.

In the following sections, we describe the inclusion of covariates into each demographic model.
We start by incrementing the intercept growth, survival, and recruitment models, as described in the previous section, with plot random effects to account for the spatial heterogeneity among plots.
Then, building on the intercept model with plot random effects, we introduce the competition for light component using individual basal area information.
Finally, we complete the model by incorporating the climate component, wherein we include the effects of mean annual temperature (MAT) and mean annual precipitation (MAP).

It's worth noting that, due to the use of structured population models, the demographic models should vary as a function of the size trait.
The von Bertalanffy growth model implicitly incorporates the size effect within the model.
The lack of available data for the regeneration process forced us to use an ingrowth rate model that is indepedent of size.
As a result, only the survival model will feature an explicit fixed-effect of size.

## Plot random effects

We account for the shared variance among individuals within the same plot in each demographic model using random effects.
In the context of each species-demographic model combination, we draw plot random effects ($\alpha_j$) from a normal distribution with a mean of zero, defined as:

$$
  \alpha_{j} \sim N(0, \sigma^2)
$$

where $\sigma^2$ represents the variance among all plots $j$.
Note that both parameters, $\alpha_j$ and $\sigma^2$, are species and demographic model-specific.
These plot random effects ($\alpha_j$) serve to adjust the intercept parameters ($I$) within each demographic model as follows:

$$
  I_j = \overline{I} + \alpha_j
$$

Where $I$ can assume one of three forms: $\Gamma$ for the growth, $\psi$ for the survival, and $\phi$ for the recruitment model.

## Size effect

For the growth model, the size effect is implicit incorporated in the model.
Specifically, following the von Bertalanffy model's definition, an individual's growth rate decreases exponentially with size, eventually reaching a zero growth rate as size approaches $\zeta_{\infty}$ (@fig-sizeEffect).

```{r,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-sizeEffect
#| fig-width: 8.5
#| fig-height: 4.5
#| fig-cap: "Illustration of the effect of size on growth rate (left panel) and survival probability (right panel) for different sets of parameters."

tibble(
    r = c(-4, -5),
    Lmax = c(1600, 1000),
    group = c(1, 2)
  ) |>
  group_by(group) |>
  expand_grid(size = seq(127, 2000, 0.1)) |>
  mutate(
    Int = Lmax * (1 - exp(-exp(r))),
    size_tp1 = Int + exp(-exp(r)) * size,
    growth = (size_tp1 - size)/size
  ) |>
  ggplot() +
  aes(size/10, growth) +
  aes(color = factor(group)) +
  geom_line() +
  geom_hline(yintercept = 0, alpha = 0.6, linetype = 3) +
  labs(
    x = 'DBH (cm)',
    y = 'Relative growth',
    color = '' 
  ) +
  scale_color_manual(
    values = met.brewer('Juarez', 2),
    labels = c(
      expression(Gamma~"="~"-4;"~zeta[infinity]~"=160"),
      expression(Gamma~"="~"-5;"~zeta[infinity]~"=100")
    )
  ) +
  theme_classic() +
  theme(legend.position = 'top') ->
p1

tibble(
    opt_size = c(400, 400, 800),
    size_var = c(10, 15, 10),
    group = c(1, 2, 3)
  ) |>
  group_by(opt_size, size_var) |>
  expand_grid(size = seq(127, 2000, 0.1)) |>
  mutate(
    sizeEffect = exp(-(log(size/opt_size)/size_var)^2)
  ) |>
  ggplot() +
  aes(size/10, sizeEffect) +
  aes(color = factor(group)) +
  geom_path() +
  geom_hline(yintercept = 1, alpha = 0.6, linetype = 3) +
  labs(
    x = 'DBH (cm)',
    y = 'Effect on survival probability',
    color = ''
  ) +
  theme_classic() +
  scale_color_manual(
    values = met.brewer('Juarez', 3),
    labels = c(
      expression(upsilon~'='~'40;'~sigma^upsilon~'='~'10'),
      expression(upsilon~'='~'40;'~sigma^upsilon~'='~'15'),
      expression(upsilon~'='~'80;'~sigma^upsilon~'='~'10')
    )
  ) +
  theme(legend.position = 'top') ->
p2

ggarrange(
  p1, p2,
  nrow = 1
)
```

In the survival model, the intercept longevity ($\psi$) varies as a log-normal function of individual dbh as follows:

$$
  \psi + \frac{ ln(\frac{DBH}{\upsilon}) }{\sigma^{\upsilon}}
$$


Here, $\upsilon$ determines the size at which survival is at its optimal and $\sigma^{\upsilon}$ quantifies the extent of survival decrease from the optimal size.
A higher parameter value corresponds to a reduced effect of size on the survival probability (@fig-sizeEffect).

## Competition effect

We use the basal area of larger individuals (BAL) as a metric for competition.
Basal area is calculated as the sum of the cross-sectional areas of all trees within a plot, derived from their diameter at breast height (dbh) measurements, and its unit is square meters per hectare.
For each focal individual, we calculate the competition intensity by summing the basal area of all individuals with a size greater than that of the focal individual. 
We differentiate this sum of basal area between conspecific and heterospecific individuals.
More details can be found in the @sec-dataset.

Both the growth ($\Gamma$) and longevity ($\psi$) intercept parameters decrease exponentially with BAL.
This negative effect of BAL on growth and longevity is driven by two parameters that describe the effect of conspecific ($\beta$) and heterospecfic ($\theta$) competition:

$$
  \Gamma + \beta^{\Gamma} \times (BAL_{cons} + \theta^{\Gamma} \times BAL_{het})
$$

$$
  \psi + \beta^{\psi} \times (BAL_{cons} + \theta^{\psi} \times BAL_{het})
$$

When $\theta < 1$, it means that conspecific competition is stronger than heterospecific competition.
Conversely, when $\theta > 1$, heterospecific competition prevails, and when $\theta = 1$, there is no distinction between conspecific and heterospecific competition (@fig-CovCompEffectGrowthMort).
Note that both $\beta$ and $\theta$ are unbounded parameters that can either converge towards negative values (indicating competition) or positive values (indivating facilitation).

```{r,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-CovCompEffectGrowthMort
#| fig-width: 8.5
#| fig-height: 4.5
#| fig-cap: "Illustration of the impact of competition on growth and survival intercept parameters. The $\\beta$ parameter governs the strength of competition, while the $\\theta$ parameter determines the extent to which this intensity is transferred to heterospecific competition. In the left panel, heterospecific competition is weaker than conspecific competition ($\\theta < 1$), whereas in the right panel, heterospecific competition is stronger ($\\theta > 1$)."

tibble(
    beta = -c(0.02, 0.05),
    theta = c(0.5, 1.5),
    group = c("1", "2")
  ) |>
  group_by(group) |>
  expand_grid(BAL = seq(0, 80, 0.1)) |>
  mutate(
    Conspecific_eff = exp(beta * BAL),
    Heterospecific_eff = exp(beta * theta * BAL)
  ) |>
  pivot_longer(cols = contains('eff')) |>
  mutate(
    name = gsub('_eff', '', name),
    group = case_match(
      group,
      "1" ~ "beta~'='~'-0.02;'~theta~'='~'0.5'",
      "2" ~ "beta~'='~'-0.05;'~theta~'='~'1.5'"
    )) |>
  ggplot() +
  aes(BAL, value) +
  aes(color = name) +
  facet_wrap(~group, labeller = label_parsed) +
  geom_path() +
  # geom_hline(yintercept = 0, alpha = 0.6, linetype = 3) +
  labs(
    x = expression('Basal area of larger individuals ('~m^2~ha^-1~')'),
    y = expression('Effect on '~Gamma~'or'~psi),
    color = '' 
  ) +
  scale_color_manual(
    values = met.brewer('Juarez', 2)
  ) +
  theme_classic() +
  theme(
    legend.position = 'top',
    strip.background = element_blank()
  )
```

For the recruitment model, conspecific and heterospecific BAL affect different components of the model.
Conspecific BAL, or simply the total conspecific plot basal area (as recruitment is necessarily smaller than any adult individual), has a unimodal effect on the annual ingrowth rate ($\phi$).
This effect is characterized by an optimal basal area for ingrowth at $\delta^{\phi}$ and an increased effect controlled by the parameter $\sigma^{\phi}$:

$$
  \phi - \left(\frac{BAL_{cons} - \delta^{\phi}}{\sigma^{\phi}}\right)^2
$$

The underlying concept of this equation is that ingrowth rate should increase with conspecific density, but only up to a certain point determined by $\delta^{\phi}$.
At higher densities, ingrowth rate is expected to decrease due to competition (@fig-CovCompEffectRec).

```{r,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-CovCompEffectRec
#| fig-width: 8.5
#| fig-height: 4.5
#| fig-cap: "Illustration of the impact of conspecific basal area on the ingrowth rate and the impact of total basal area on the survival rate of recruited individuals."

tibble(
    opt_BA = c(5, 40, 40),
    sigma_BA = c(20, 20, 60),
    group = c('A', 'B', 'C')
  ) |>
  group_by(group) |>
  expand_grid(BAL = seq(0, 80, 0.1)) |>
  mutate(
    cons_eff = exp(-((BAL - opt_BA)/sigma_BA)^2)
  ) |>
  ggplot() +
  aes(BAL, cons_eff) +
  aes(color = group) +
  geom_path() +
  labs(
    x = expression('BAL'['cons']~'('~m^2~ha^-1~')'),
    y = expression('Effect on '~phi),
    color = '' 
  ) +
  scale_color_manual(
    values = met.brewer('Juarez', 3),
    labels = c(
      expression(delta^phi~'='~'5;'~sigma^phi~'='~'10'),
      expression(delta^phi~'='~'40;'~sigma^phi~'='~'10'),
      expression(delta^phi~'='~'40;'~sigma^phi~'='~'30')
    )
  ) +
  theme_classic() +
  theme(
    legend.position = 'top'
  ) ->
p1

tibble(
    beta = -c(0.02, 0.05, 0.08),
    group = c('A', 'B', 'C')
  ) |>
  group_by(group) |>
  expand_grid(BAL = seq(0, 80, 0.1)) |>
  mutate(
    cons_eff = exp(beta * BAL)
  ) |>
  ggplot() +
  aes(BAL, cons_eff) +
  aes(color = group) +
  geom_path() +
  labs(
    x = expression('BAL'['cons']~'+'~'BAL'['het']~'('~m^2~ha^-1~')'),
    y = expression('Effect on '~rho),
    color = '' 
  ) +
  scale_color_manual(
    values = met.brewer('Juarez', 3),
    labels = c(
      expression(beta^rho~'='~'-0.02'),
      expression(beta^rho~'='~'-0.05'),
      expression(beta^rho~'='~'-0.08')
    )
  ) +
  theme_classic() +
  theme(
    legend.position = 'top'
  ) ->
p2

ggarrange(
  p1, p2,
  nrow = 1
)
```

Finally, the annual survival probability reduces exponentially with the total basal area of the plot, where there is no distinction between conspecific and heterospecific competition:

$$
  \rho + \beta^{\rho} \times (BAL_{cons} +  BAL_{het})
$$


## Climate effect

As our focus is on inference rather than prediction, we chose to not conduct model selection concerning the climate variables.
Instead, we opted for the mean annual temperature (MAT) and mean annual precipitation (MAP) as our chosen bioclimatic variables, which are widely used for species distribution modeling.
Each demographic function varies in function of a bell-shaped curve determined by an optimal climate condition ($\xi$) and a climate breadth parameter ($\sigma$) as follows:

$$
  \Gamma + \left(\frac{MAT - \xi^{\Gamma}_{MAT}}{\sigma^{\Gamma}_{MAT}}\right)^2 + \left(\frac{MAP - \xi^{\Gamma}_{MAP}}{\sigma^{\Gamma}_{MA{P}}}\right)^2
$$

$$
  \psi + \left(\frac{MAT - \xi^{\psi}_{MAT}}{\sigma^{\psi}_{MAT}}\right)^2 + \left(\frac{MAP - \xi^{\psi}_{MAP}}{\sigma^{\psi}_{MA{P}}}\right)^2
$$

$$
  \phi + \left(\frac{MAT - \xi^{\phi}_{MAT}}{\sigma^{\phi}_{MAT}}\right)^2 + \left(\frac{MAP - \xi^{\phi}_{MAP}}{\sigma^{\phi}_{MA{P}}}\right)^2
$$

The climate breadth parameter ($\sigma$) influences the strength of the specific climate variable's effect on each demographic model (@fig-CovClim).
This unimodal function is flexible as it can assume various shapes to better accommodate the data.
However, this flexibility also introduces the possibility of parameter degeneracy or redundancy, where different combinations of parameter values yield similar outcomes.
To mitigate parameter degeneracy, we constrained the optimal climate condition parameter ($\xi$) within the observed climate range for the species, thereby assuming that the optimal climate condition falls within our data range.

```{r,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-CovClim
#| fig-width: 5.5
#| fig-height: 4
#| fig-cap: "Illustration of a hypothetical scaled climate variable's impact on the growth, survival, and recruitment intercept parameters."

tibble(
    opt_clim = c(0, 0.5, 0.5, 1),
    sigma_clim = c(.5, .3, 1, .5),
    group = c('A', 'B', 'C', 'D')
  ) |>
  group_by(group) |>
  expand_grid(clim = seq(0, 1, 0.01)) |>
  mutate(
    clim_eff = exp(-((clim - opt_clim)/sigma_clim)^2)
  ) |>
  ggplot() +
  aes(clim, clim_eff) +
  aes(color = group) +
  geom_path() +
  labs(
    x = 'Climate variable (scaled)',
    y = expression('Effect on '~Gamma~'/'~psi~'/'~phi),
    color = '' 
  ) +
  scale_color_manual(
    values = met.brewer('Juarez', 4),
    labels = c(
      expression(xi~'='~'0;'~sigma~'='~'0.5'),
      expression(xi~'='~'0.5;'~sigma~'='~'0.3'),
      expression(xi~'='~'0.5;'~sigma~'='~'1'),
      expression(xi~'='~'1;'~sigma~'='~'0.5')
    )
  ) +
  theme_classic() +
  theme(
    legend.position = 'top'
  )
```
