# Intercept {#sec-parsIntercept}

Distribution of intercept for the growth, survival, and recruitment models.

```{r,include=FALSE,echo=FALSE}
Echo=FALSE
Eval=TRUE
Cache=TRUE
Warng=FALSE
Msg=FALSE
library(tidyverse)
library(cmdstanr)
library(posterior)
library(ggdist)
library(ggpubr)
library(ggrepel)
```


```{r intercept_loadPars,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}

# Using parameters from the `forest-IPM` repo
spIds <- read_csv(
  file.path(
    '..', 'data', 'species_id.csv'
  )
) |>
mutate(
  shade = factor(shade, levels = c('tolerant', 'intermediate', 'intolerant')),
  shade_sylvics = factor(shade_sylvics, levels = c('very-tolerant',
                                            'tolerant',
                                            'intermediate',
                                            'intolerant',
                                            'very-intolerant'
                                            )),
  succession = factor(succession, levels = c('pioneer',
                                              'intermediate',
                                              'climax'))
) |>
filter(sp_to_analyze)

map_dfr(
  spIds$species_id_old,
  ~ read_csv(
    paste0(
      '../data/parameters/growth_', .x, '.csv'
    ),
    show_col_types = FALSE
  ) |>
  select(c(r, Lmax)) |>
  bind_cols(species_id = .x)
) ->
pars_growth

map_dfr(
  spIds$species_id_old,
  ~ read_csv(
    paste0(
      '../data/parameters/mort_', .x, '.csv'
    ),
    show_col_types = FALSE
  ) |>
  select('psi') |>
  bind_cols(species_id = .x)
) ->
pars_mort

map_dfr(
  spIds$species_id_old,
  ~ read_csv(
    paste0(
      '../data/parameters/rec_', .x, '.csv'
    ),
    show_col_types = FALSE
  ) |>
  select(c(mPop_log, p_log)) |>
  bind_cols(species_id = .x)
) ->
pars_rec

treeData <- readRDS(file.path('..', 'data', 'treeData.RDS')) |>
  filter(species_id %in% spIds$species_id_old)

```


## Growth rate

Because the growth rate intercept decreases non-linearly with size and is governed by the interaction of two parameters ($r$ and $S_{\infty}$), we computed the average growth rate over a 10 years interval.
Also, because the size of the individual is integrated in the Von Bertalanffy growth model, we computed the 10-years growth average starting from the lower size threshold of 12.7 cm, where growth is at its optimal.

```{r intcerpt_growth,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-intGrowth
#| fig-width: 8.5
#| fig-height: 5.5
#| fig-cap: "Intercept of the growth model using the 10 years average growth rate. Species are classified by their general growth trait following @burns1990silvics"

pars_growth |>
  mutate(
    size_t0 = 127,
    deltaTime = 10,
    rPlotTime = exp(-exp(r) * deltaTime),
    size_t1 = size_t0 * rPlotTime + Lmax * (1 - rPlotTime),
    growth = (size_t1 - size_t0)/deltaTime
  ) |>
  # # This is the formula for the intercept (growth at zero close to zero)
  # mutate(
  #   Int = Lmax * exp(-exp(r))
  # ) |>
  left_join(
    spIds,
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(growth, fct_reorder(species_name, growth)) +
  aes(color = growth_rate) +
  stat_pointinterval() +
  ylab('') +
  xlab('10 years average growth rate (mm/year)') +
  scale_color_manual(
    values = c("#87bc45", "#edbf33", "#ea5545")
  ) +
  theme_classic() +
  theme(
    axis.text.y = element_text(face = "italic"),
    legend.position = 'none'
  ) ->
p1

pars_growth |>
  mutate(
    size_t0 = 127,
    deltaTime = 10,
    rPlotTime = exp(-exp(r) * deltaTime),
    size_t1 = size_t0 * rPlotTime + Lmax * (1 - rPlotTime),
    growth = (size_t1 - size_t0)/deltaTime
  ) |>
  left_join(
    spIds,
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(growth_rate, growth) +
  aes(fill = growth_rate) +
  geom_boxplot() +
  scale_fill_manual(
    values = c("#87bc45", "#edbf33", "#ea5545")
    # values = c("#20bc45", "#87bc45", "#edbf33", "#ea5545", "#ba0000")
  ) +
  xlab('') +
  ylab('Average 10-years growth rate (mm/year)') +
  theme_classic() +
  theme(legend.position = 'none') +
  labs(fill = '') ->
p2

print(ggarrange(p1, p2, ncol = 2))
```

In the following figure we compare the maximum observed size in the literature with the asymptotic size ($S_{\infty}$) denoting the size in which growth rate converge to zero.

```{r intcerpt_maxObsLmax,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-efszGrowth2
#| fig-width: 6.5
#| fig-height: 5
#| fig-cap: "Correlation between maximum observed size in literature following @burns1990silvics and estimated parameter $S_{\\infty}$"

parsGrowth_mean <- pars_growth |>
  group_by(species_id) |>
  reframe(
    Lmax = mean(Lmax),
    r = mean(r)
  ) |>
  left_join(
    spIds,
    by = c('species_id' = 'species_id_old')
  ) 

pars_growth |>
  left_join(
    spIds,
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(max_size/10, Lmax/10, group = species_id) +
  # aes(color = shade_sylvics) +
  stat_pointinterval(alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, alpha = 0.3) +
  geom_text_repel(
    data = parsGrowth_mean,
    aes(x = max_size/10, y = Lmax/10, label = species_name),
    alpha = 0.8,
    size = 2.6,
    fontface = 'italic'
  ) +
  theme_classic() +
  xlab('Maximum observed dbh (cm)') +
  ylab('Maximum predicted dbh (cm)')
```


## Mortality probability

Similar to the growth rate, the surival model has a temporal component in which survival probability ($logit~\psi$) decreases exponently with time.
So we computed the 10 years average mortality probability as the intercept for the survival model.

```{r intcerpt_survival,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-efszMort
#| fig-width: 8.5
#| fig-height: 5.5
#| fig-cap: "Intercept of the survival model using a 10 years mortality probability. Species are classified by their shade tolerance trait following @burns1990silvics"

pars_mort |>
  mutate(psi = 1 - 1/(1+exp(-psi))^10) |>
  left_join(
    spIds,
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(psi, fct_reorder(species_name, psi)) +
  aes(color = shade_sylvics) +
  stat_pointinterval() +
  ylab('') +
  xlab('10 years mortality probability') +
  scale_color_manual(
    # values = c("#87bc45", "#edbf33", "#ea5545")
    values = c("#20bc45", "#87bc45", "#edbf33", "#ea5545", "#ba0000")
  ) +
  theme_classic() +
  theme(
    axis.text.y = element_text(face = "italic"),
    legend.position = 'none'
  ) ->
p1

pars_mort |>
  mutate(psi = 1 - (1/(1+exp(-psi))^10)) |>
  left_join(
    spIds,
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(shade_sylvics, psi, fill = shade_sylvics) +
  geom_boxplot() +
  scale_fill_manual(
    # values = c("#87bc45", "#edbf33", "#ea5545")
    values = c("#20bc45", "#87bc45", "#edbf33", "#ea5545", "#ba0000")
  ) +
  xlab('Shade tolerance') +
  ylab('10 years mortality probability') +
  theme_classic() +
  theme(legend.position = 'none') +
  labs(fill = '') ->
p2

ggarrange(p1, p2, ncol = 2)
```

Alternatively, we can also use the parameter $\psi$ to derive the expected longevity ($L$) which can be determined as $L = e^{\psi}$.
In @fig-efszMort2, we plot the correlation between maximum size observed in the literature [@burns1990silvics] with longevity ($L$).

```{r intcerpt_maxAgevsPsi,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-efszMort2
#| fig-width: 7.5
#| fig-height: 5
#| fig-cap: "Correlation between maximum size observed in the literature [@burns1990silvics] and the predicted longevity posterior distribution."

parsMort_mean <- pars_mort |>
  group_by(species_id) |>
  reframe(psi = mean(psi)) |>
  left_join(
    spIds,
    by = c('species_id' = 'species_id_old')
  ) 

pars_mort |>
  left_join(
    spIds,
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(max_age, exp(psi), group = species_id) +
  # aes(color = shade_sylvics) +
  stat_pointinterval(alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, alpha = 0.3) +
  geom_text_repel(
    data = parsMort_mean,
    aes(x = max_age, y = exp(psi), label = species_name),
    alpha = 0.8,
    size = 2.6,
    fontface = 'italic'
  ) +
  theme_classic() +
  xlab('Maximum observed age') +
  ylab('Expected longevity')
```


## Ingrowth rate

The ingrowth model comprises two intercept components: the annual ingrowthrate per square meter ($m$) and the annual survival probability $p$.
In the context of the survival submodel for the ingrowth rate, we computed the 10 years survival probability of ingrowth individuals just like in the survival model.

```{r intcerpt_ingrowth,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| fig-width: 8.5
#| fig-height: 5.5
#| fig-cap: "Intercept of the ingrwoth model for the number of individuals that ingress the population per year per m$^2$. Species are classified by their successional status."

# Fig 3: m ingrowth in function of species and shade tolerance
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
  pars_rec |>
    filter(mPop_log < quantile(mPop_log, 0.975)) |>
    left_join(
      spIds,
      by = c('species_id' = 'species_id_old')
    ) |>
    ggplot() +
    aes(exp(mPop_log), fct_reorder(species_name, exp(mPop_log))) +
    aes(color = succession) +
    stat_pointinterval() +
    ylab('') +
    xlab('Annual ingrowth rate per m²') +
    scale_color_manual(
      values = c('#f7fcb9', '#addd8e', '#31a354')
      # values = c("#87bc45", "#edbf33", "#ea5545")
      # values = c("#20bc45", "#87bc45", "#edbf33", "#ea5545", "#ba0000")
    ) +
    theme_classic() +
    theme(
      axis.text.y = element_text(face = "italic"),
      legend.position = 'none'
    ) ->
  p1

  pars_rec |>
    filter(mPop_log < quantile(mPop_log, 0.975)) |>
    left_join(
      spIds,
      by = c('species_id' = 'species_id_old')
    ) |>
    filter(!is.na(succession)) |>
    ggplot() +
    aes(log(sm), exp(mPop_log), color = succession) +
    stat_pointinterval() +
    scale_color_manual(
      values = c('#f7fcb9', '#addd8e', '#31a354')
      # values = c("#87bc45", "#edbf33", "#ea5545")
      # values = c("#20bc45", "#87bc45", "#edbf33", "#ea5545", "#ba0000")
    ) +
    theme_classic() +
    labs(
      x = 'Seed mass (ln g)',
      y = 'Annual ingrowth rate per m²',
      color = ''
    ) +
    theme(legend.position = 'top') ->
  p2

  ggarrange(p1, p2, ncol = 2)
```


```{r intcerpt_ingP,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| fig-width: 8.5
#| fig-height: 5.5
#| fig-cap: "Intercept of the annual survival probability for the ingrwoth model. Species are classified by their shade tolerance trait following @burns1990silvics"

pars_rec |>
  mutate(p = exp(-exp(p_log))^10) |>
  left_join(
    spIds,
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(p, fct_reorder(species_name, p)) +
  aes(color = shade_sylvics) +
  stat_pointinterval() +
  ylab('') +
  xlab('10 years survival probability') +
  scale_color_manual(
    # values = c("#87bc45", "#edbf33", "#ea5545")
    values = c("#20bc45", "#87bc45", "#edbf33", "#ea5545", "#ba0000")
  ) +
  theme_classic() +
  theme(
    axis.text.y = element_text(face = "italic"),
    legend.position = 'none'
  ) ->
p1

pars_rec |>
  left_join(
    spIds,
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(shade_sylvics, exp(-exp(p_log))^10) +
  aes(fill = shade_sylvics) +
  geom_boxplot() +
  scale_fill_manual(
    # values = c("#87bc45", "#edbf33", "#ea5545")
    values = c("#20bc45", "#87bc45", "#edbf33", "#ea5545", "#ba0000")
  ) +
  ylab('10 years survival probability') +
  xlab('') +
  theme_classic() +
  theme(legend.position = 'none') ->
p2

ggarrange(p1, p2, ncol = 2)
```
