# Competition effect {#sec-parsComp}

In this section I discuss the competition effect on the growth, survival, and recruitment vital rates.
For all vital rates, competition for light increases linearly with density at the plot level following a general $\beta$ parameter meaning the density dependence effect.
As we have the size measurement of each adult individual for the growth and survival models, we used the total basal area of individuals larger than the focal individual as a competition index.
Furthermore, we split the total basal area effect between conspecific and heterospecific individuals larger than the focal individual.
As the recruitment model does not have individual information, we simply used the total basal area of all alive adults.

```{r,include=FALSE,echo=FALSE}
Echo=FALSE
Eval=TRUE
Cache=TRUE
Warng=FALSE
Msg=FALSE
library(tidyverse)
library(cmdstanr)
library(posterior)
library(ggdist)
library(ggpubr)
library(ggrepel)
```

```{r comp_loadPars,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
data_path <- readLines('_data.path')
pars_path <- file.path(data_path, 'output_sim_processed')
models <- c(
  'growth' = 'intcpt_plot_comp_clim',
  'mort' = 'intcpt_plot_comp_clim',
  'recruit' = 'intcpt_plot_comp_clim'
)

## Species if info
spIds <- read_csv(
  file.path(
    data_path, 'species_id.csv'
  )
) |>
mutate(
  shade = factor(shade, levels = c('tolerant', 'intermediate', 'intolerant')),
  shade_sylvics = factor(shade_sylvics, levels = c('very-tolerant',
                                            'tolerant',
                                            'intermediate',
                                            'intolerant',
                                            'very-intolerant'
                                            )),
  succession = factor(succession, levels = c('pioneer',
                                              'intermediate',
                                              'climax'))
) |>
filter(sp_to_analyze)


# Load parameters
map_dfr(
  spIds$species_id_old,
  ~ readRDS(
    paste0(
      pars_path, '/growth/', models['growth'], '/posterior_pop_', .x, '.RDS'
    )
  ) |>
  filter(par %in% c('Beta', 'theta')) |>
  pivot_wider(names_from = par) |>
  bind_cols(species_id = .x)
) ->
pars_growth

map_dfr(
  spIds$species_id_old,
  ~ readRDS(
    paste0(
      pars_path, '/mort/', models['mort'], '/posterior_pop_', .x, '.RDS'
    )
  ) |>
  filter(par %in% c('Beta', 'theta')) |>
  pivot_wider(names_from = par) |>
  bind_cols(species_id = .x)
) ->
pars_survival

map_dfr(
  spIds$species_id_old,
  ~ readRDS(
    paste0(
      pars_path, '/recruit/', models['recruit'], '/posterior_pop_', .x, '.RDS'
    )
  ) |>
  filter(par %in% c('beta_p', 'optimal_BA', 'sigma_BA')) |>
  pivot_wider(names_from = par) |>
  bind_cols(species_id = .x)
) ->
pars_rec

treeData <- readRDS(file.path(data_path, 'treeData.RDS')) |>
  filter(species_id %in% spIds$species_id_old)
```


## Conspecific and heterospecific effect

Using the basal area of conspecific and heterospecific larger individuals than the focal individual.

#### Growth

```{r comp_compGrowth,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-compGrowth
#| fig-width: 8.5
#| fig-height: 6
#| fig-cap: "Posterior distribution of conspecific and heterospecific density dependence across species (left panel) and species shade tolerance (right panel). Species are classified by their general growth trait following @burns1990silvics"

pars_growth |>
  mutate(
    Con = Beta,
    Het = Beta * theta
  ) |>
  pivot_longer(cols = c(Con, Het)) |>
  left_join(
    spIds,
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(value, fct_reorder(species_name, value)) +
  aes(color = name) +
  stat_pointinterval() +
  geom_vline(xintercept = 0, alpha = 0.7) +
  ylab('') +
  xlab('Density dependence effect') +
  # scale_color_manual(
  #   values = c('#5ab4ac', '#d8b365')
  # ) +
  theme_classic() +
  theme(
    axis.text.y = element_text(face = "italic"),
    legend.position = 'none'
  ) +
  labs(color = '') ->
p1

pars_growth |>
  mutate(
    Con = Beta,
    Het = Beta * theta
  ) |>
  pivot_longer(cols = c(Con, Het)) |>
  left_join(
    spIds,
    by = c('species_id' = 'species_id_old')
  ) |>
  mutate(
    name = case_match(
      name,
      'Con' ~ 'Conspecific',
      'Het' ~ 'Heterospecific'
    )
  ) |>
  ggplot() +
  aes(shade_sylvics, value) +
  aes(fill = name) +
  geom_boxplot() +
  scale_color_manual(
    values = c('#5ab4ac', '#d8b365')
  ) +
  xlab('') +
  ylab('Density dependence effect') +
  theme_classic() +
  theme(legend.position = 'top') +
  labs(fill = '') ->
p2

ggarrange(p1, p2, ncol = 2)
```


#### Survival

```{r comp_compSurv,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-compSurvival
#| fig-width: 8.5
#| fig-height: 6
#| fig-cap: "Posterior distribution of conspecific and heterospecific density dependence across species (left panel) and species shade tolerance (right panel). Species are classified by their general growth trait following @burns1990silvics"

pars_survival |>
  mutate(
    Con = Beta,
    Het = Beta * theta
  ) |>
  pivot_longer(cols = c(Con, Het)) |>
  left_join(
    spIds,
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(value, fct_reorder(species_name, value)) +
  aes(color = name) +
  stat_pointinterval() +
  geom_vline(xintercept = 0, alpha = 0.7) +
  ylab('') +
  xlab('Density dependence effect') +
  # scale_color_manual(
  #   values = c("#87bc45", "#edbf33", "#ea5545")
  # ) +
  theme_classic() +
  theme(
    axis.text.y = element_text(face = "italic"),
    legend.position = 'none'
  ) +
  labs(color = '') ->
p1


pars_survival |>
  mutate(
    Con = Beta,
    Het = Beta * theta
  ) |>
  pivot_longer(cols = c(Con, Het)) |>
  left_join(
    spIds,
    by = c('species_id' = 'species_id_old')
  ) |>
  mutate(
    name = case_match(
      name,
      'Con' ~ 'Conspecific',
      'Het' ~ 'Heterospecific'
    )
  ) |>
  ggplot() +
  aes(shade_sylvics, value) +
  aes(fill = name) +
  geom_boxplot() +
  # scale_fill_manual(
  #   values = c("#87bc45", "#edbf33", "#ea5545")
  #   # values = c("#20bc45", "#87bc45", "#edbf33", "#ea5545", "#ba0000")
  # ) +
  xlab('') +
  ylab('Density dependence effect') +
  theme_classic() +
  theme(legend.position = 'top') +
  labs(fill = '') ->
p2

ggarrange(p1, p2, ncol = 2)
```


#### Growth vs survival

```{r comp_compGrowthVsSurv,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-compGrowthVsSurv
#| fig-width: 8.5
#| fig-height: 5
#| fig-cap: "Conspecific and heterospecific density dependence between growth and survival."

pars_growth |>
  mutate(
    Con_growth = Beta,
    Het_growth = Beta * theta
  ) |>
  select(!c(Beta, theta)) |>
  group_by(species_id) |>
  mutate(iter = 1:n()) |>
  left_join(
    pars_survival |>
      mutate(
        Con_mort = Beta,
        Het_mort = Beta * theta
      ) |>
      select(!c(Beta, theta)) |>
      group_by(species_id) |>
      mutate(iter = 1:n())
  ) |>
  pivot_longer(
    cols = c(contains('Het_'), contains('Con_')),
    names_to = c('comp', 'vr'),
    names_sep = '_'
  ) |>
  mutate(
    comp = case_match(
      comp,
      'Con' ~ 'Conspecific',
      'Het' ~ 'Heterospecific'
    )
  ) |>
  pivot_wider(
    names_from = vr,
    values_from = value
  ) |>
  left_join(
    spIds,
    by = c('species_id' = 'species_id_old')
  ) |>
  group_by(species_id, comp) |>
  slice_sample(n = 1000) |>
  ggplot() +
  aes(growth, mort) +
  # aes(color = shade_sylvics) +
  facet_wrap(~comp, scales = 'free') +
  geom_point(alpha = 0.1, size = 0.1) +
  geom_abline(intercept = 0, slope = 1, alpha = 0.5) +
  xlab('Growth density dependence') +
  ylab('Survival density dependence') +
  scale_color_manual(
    # values = c("#87bc45", "#edbf33", "#ea5545")
    values = c("#20bc45", "#87bc45", "#edbf33", "#ea5545", "#ba0000")
  ) +
  theme_classic() +
  labs(color = '')
```


## Total basal area effect on recruitment survival

```{r comp_compGrowthVsSurv,echo=Echo,eval=Eval,cache=Cache,warning=Warng,message=Msg}
#| label: fig-compRec
#| fig-width: 8.5
#| fig-height: 6
#| fig-cap: "Posterior distribution of the density depndence parameter affecting annual survival rate of recruitment individuals."

pars_rec |>
  left_join(
    spIds,
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(-beta_p, fct_reorder(species_name, -beta_p)) +
  # aes(color = ) +
  stat_pointinterval() +
  geom_vline(xintercept = 0, alpha = 0.7) +
  ylab('') +
  xlab('Density dependence effect') +
  # scale_color_manual(
  #   values = c("#87bc45", "#edbf33", "#ea5545")
  # ) +
  theme_classic() +
  theme(
    axis.text.y = element_text(face = "italic"),
    legend.position = 'none'
  ) +
  labs(color = '') ->
p1


pars_rec |>
  left_join(
    spIds,
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(shade_sylvics, -beta_p) +
  aes(fill = shade_sylvics) +
  geom_boxplot() +
  scale_fill_manual(
    # values = c("#87bc45", "#edbf33", "#ea5545")
    values = c("#20bc45", "#87bc45", "#edbf33", "#ea5545", "#ba0000")
  ) +
  xlab('') +
  ylab('Density dependence effect') +
  theme_classic() +
  theme(legend.position = 'none') +
  labs(fill = '') ->
p2

ggarrange(p1, p2, ncol = 2)
```
